---
title: "Read-based model for swab P2RA"
format:
  html:
    code-fold: true
    toc: true
  pdf:
    toc: true
knitr:
  opts_chunk: 
    message: false
editor: source 
author: dan
date: 2025-05-27
---

# Introduction

This document provides a statistical analysis approach for estimating virus prevalence from pooled nasal swab samples, and for calculating the relative abundance at 1% prevalence (RA_p(1%)) from wastewater metagenomic sequencing data. The approach addresses concerns raised in the internal memo regarding the sensitivity of detection in pooled samples with low read counts.

The key concerns being addressed include:

1. The possibility that low viral read counts in positive pools indicate that we're missing many true positive samples

2. The challenge of distinguishing between scenarios where we're detecting most positive samples versus scenarios where we're only detecting a fraction of them

3. The need for a statistical model that accounts for imperfect detection sensitivity

This analysis uses a Bayesian approach to jointly model both the prevalence data from swab samples and the relative abundance data from wastewater samples, allowing for uncertainty in detection.

## Setup

First, we'll load the necessary libraries and set up our analysis environment:

```{r setup}
library(tidyverse)
library(rstan)
library(knitr)
library(kableExtra)
theme_set(theme_minimal())

# Colors borrowed from https://github.com/JLSteenwyk/ggpubfigs
wong_eight <- c(
  "#E69F00",
  "#56B4E9",
  "#009E73",
  "#F0E442",
  "#0072B2",
  "#D55E00",
  "#CC79A7",
  "#000000"
)
options(
  ggplot2.discrete.colour = function()
    scale_colour_manual(values = wong_eight)
)
options(
  ggplot2.discrete.fill = function()
    scale_fill_manual(values = wong_eight)
)

options(mc.cores = parallel::detectCores())

# For axis labels
scientific <- function(x) {
  parse(text = sprintf("10^{%f}", log10(x)))
}
```

# Data 

Now we'll load and prepare the data for our analysis:

```{r}
#| code-fold: false
# Swab data

swab_metadata <- read_tsv("../../tables/swab-sample-metadata.tsv") %>%
  mutate(date = ymd(paste0("20", date)))

# Use the per-day data. Caveat about the different treatments
swab_reads <- read_tsv("../../tables/swabs-ra-summary.tsv") %>%
  mutate(date = ymd(paste0("20", date))) %>%
  select(-all_reads)

# Wastewater data

wastewater_metadata <- read_tsv("../../tables/wastewater-sample-metadata.tsv") %>%
  mutate(date = ymd(paste0("20", date)))

wastewater_reads <- read_tsv("../../tables/ww-ra-summary.tsv") %>%
  mutate(date = ymd(paste0("20", date))) %>%
  select(-all_reads)
```

Next, we'll organize the virus species data and prepare it for analysis.
We'll look at all species found in either dataset.

```{r}
#| code-fold: false

all_species <- bind_rows(swab_reads, wastewater_reads) %>%
  select(species, group) %>%
  distinct %>%
  arrange(group, species)

species_order <- all_species$species

# Fill in zero counts
swab_reads_complete <- crossing(all_species, swab_metadata) %>%
  left_join(
    summarise(
      swab_reads,
      .by = c(date, location, species, group),
      viral_reads = sum(dedup_hv)
    ),
    by = join_by(date, location, species, group),
  ) %>%
  arrange(group, species) %>%
  mutate(
    viral_reads = replace_na(viral_reads, 0),
    species = fct_inorder(species),
    group = fct_inorder(group)
  )

wastewater_reads_complete <- wastewater_metadata %>%
  crossing(all_species) %>%
  left_join(
    summarise(
      wastewater_reads,
      .by = c(date, location, species, group),
      viral_reads = sum(dedup_hv)
    ),
    by = join_by(date, location, species, group),
  ) %>%
  arrange(group, species) %>%
  mutate(
    viral_reads = replace_na(viral_reads, 0),
    species = fct_inorder(species),
    group = fct_inorder(group)
  )
```

This data preparation ensures we have complete data for all species across all samples, filling in zeros where necessary. We organize the data by virus species and group for easier analysis.

```{r}
species_swab_summary <- swab_reads_complete %>%
  summarise(
    pos_pools = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    .by = c(species, group),
  ) %>%
  arrange(species)
kable(species_swab_summary)
```

```{r}
species_ww_summary <- wastewater_reads_complete %>%
  summarise(
    pos_samples = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    .by = c(species, group),
  ) %>%
  arrange(species)
kable(species_ww_summary)
```

# Model

Our goal is to jointly estimate virus prevalence from pooled swab samples and relative abundance from wastewater samples, while:

1.  **Accounting for imperfect detection**: The model doesn't assume that all positive individuals will be detected. Instead, it models the probability of detecting various numbers of viral reads given the number of positive individuals in a pool.

2.  **Modeling the viral fraction directly**: The parameter $\mu_{swab}$ represents the expected viral fraction in a pool with only positive swabs. This can be very small, which would indicate low sensitivity.

3.  **Integrating over uncertainty**: Rather than assuming we know how many positive individuals are in each pool, the model integrates over all possibilities, weighted by their probability given the observed data.

4.  **Joint modeling with wastewater data**: By modeling both swab and wastewater data together, we can more naturally incorporate different sources of uncertainty.

## Parameters

-   $p$: Prevalence of the virus in the population (probability an individual is infected)
-   $\mu_{swab}$: Expected viral fraction in a pool containing only positive swabs.
    The expected viral fraction in a pool drawn from a population with prevalence $p$ is $\mu_{swab} p$.
-   $\phi_1$: Inverse-overdispersion parameter for viral reads in a pool in which a single swab is positive
-   $\mu_{ww}$: Expected viral fraction in wastewater if everyone were infected.
    The expected viral fraction in wastewater collected from a population with prevalence $p$ is $\mu_{ww} p$.
-   $\phi_{ww}$: Inverse-overdispersion parameter for viral reads in wastewater

## Swab Sample Model

For each pool $i$ with $s_{i}$ swabs:

1.  The number of positive swabs $s_{+,i}$ follows a binomial distribution: $$s_{+,i} \sim \text{Binomial}(s_{i}, p)$$

2.  The viral fraction $v_i$ in the pool depends on the number of positive swabs: $$v_i \sim \text{Gamma}\left(s_{+,i}\phi_1, \frac{s_{i}\phi_1}{\mu_{swab}}\right)$$
    (in shape-rate parameterization)
    This gives: $$E[v_i] = \frac{s_{+,i}}{s_{i}} \cdot \mu_{swab}$$ $$\text{Var}[v_i] = \frac{s_{+,i}}{s_{i}^2} \cdot \frac{\mu_{swab}^2}{\phi_1}$$

3.  The viral read count follows a Poisson distribution conditional on the viral fraction: $$r_{viral,i} \mid v_i \sim \text{Poisson}(r_{total,i} \cdot v_i)$$

4.  Marginalizing over $v_i$, the viral read count follows a negative binomial distribution: $$r_{viral,i} \mid s_{+,i} \sim \text{NegBinom2}\left(\frac{s_{+,i} \cdot r_{total,i} \cdot \mu_{swab}}{s_{i}}, s_{+,i} \cdot \phi_1\right)$$

5.  Since we don't observe $s_{+,i}$, the likelihood for a pool is: $$P(r_{viral,i}) = \sum_{s_{+,i}=0}^{s_{i}} P(r_{viral,i} \mid s_{+,i}) \cdot P(s_{+,i})$$

    Where:

    -   If $s_{+,i} = 0$, then $P(r_{viral,i} \mid s_{+,i} = 0) = \mathbf{1}[r_{viral,i} = 0]$
    -   Otherwise, $P(r_{viral,i} \mid s_{+,i})$ is given by the negative binomial above
    -   $P(s_{+,i})$ is given by the binomial distribution with parameters $s_{i}$ and $p$

## Wastewater Sample Model

For each wastewater sample $j$:

1. The expected viral read count is proportional to the prevalence: $$E[r_{viral_j}] = r_{total,j} \cdot p \cdot \mu_{ww}$$

2. The viral read count follows a negative binomial distribution: $$r_{viral,j} \sim \text{NegBinom2}(r_{total,j} \cdot p \cdot \mu_{ww}, \phi_{ww})$$

## Prior Distributions

### Prevalence

Previously we used a $\text{Beta}(0.5, 10)$ prior but this may be too constraining on the posteriors.

Let's try to reason about plausible upper and lower bounds for $p$:

- Because we see all of these viruses in our wastewater sequencing, we can probably assume 
that at least one person in Boston was infected and shedding.
There are 5 million people in Greater Boston, so that suggests a lower bound of ~2e-7.
(There are fewer people living in the DITP sewershed but this accounts for commuters, etc.)

- On the other end, consider a virus that everyone gets over the course of a three-month outbreak
and that detectable shedding lasts for a month. Then something like half of people will be contribute
to the effective prevalence at its peak.

With these as guides, 
We'll put a logit-Normal prior on $p$, which constrains the values of $p$ to be between zero and one:
$$\text{logit}(p) \sim \text{Normal}(-6, 2)$$
```{r}
#| label: prior-p

mean_logit_p <- -6
sd_logit_p <- 2
prior_logit_p <- function(x) dnorm(x, mean_logit_p, sd_logit_p)

tibble(
  logit_p = seq(-20, 10, length.out = 1000),
  p = plogis(logit_p),
  prior = prior_logit_p(logit_p),
) %>%
  ggplot(aes(p, prior)) +
  geom_line() +
  geom_vline(xintercept = plogis(mean_logit_p), linetype = "solid", color = "grey") +
  geom_vline(xintercept = plogis(mean_logit_p + c(-1, 1) * sd_logit_p), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = c(0.5), linetype = "dotted", color = "blue") +
  geom_vline(xintercept = c(2e-7), linetype = "dotted", color = "red") +
  scale_x_continuous(
    transform = "logit",
    breaks = c(10^seq(-8, -1, 2), 0.5, 1 - 10^seq(-2, -1, 2)),
    name = "p (logistic scale)",
  ) +
  labs(title = "Prior on prevalence", y = "")
```

This is a very broad prior.
It implies 4% probability on more than half of people being positive (blue dotted line)
and 2% of fewer than one person in Greater Boston being positive in expectation (red dotted line).

The median $p$ is at $\text{logit}^{-1}(-7) = 9 \times 10^{-4}$ (solid grey line).
For comparison, there were 547 swabs, so one positive swab would have a positive fraction of 1.8e-3.

About 70% of the prior probability is between 1.6e-5 and 0.047 (grey dashed lines).

This prior thus assumes that the virus is probably present in Boston,
but does not assume that it’s at a prevalence that’s detectable in the swab data.
The current analysis does not include the viruses that were detected in wastewater
but not swabs, but if we added them back in, we would want a prior that takes the possibility
of very low prevalence seriously.

### Expected viral fractions

Note that $\mu_{swab}$ primarily matters to our estimate of $p$ through the probability that
a positive swab contributes zero viral reads in a pool.
Lower $\mu_{swab}$ implies greater probability that a positive swab does not contribute viral reads,
which in turn implies higher prevalence.
Higher prevalence shifts our estimate of $\mu_{ww}$ downward.

For the viral fraction parameters ($\mu_{swab}$ and $\mu_{ww}$), we use logit-normal priors to constrain these values between 0 and 1.

Because we do not have much prior information about these parameters,
we chose priors that span many orders of magnitude.
Our previous estimates for Flu and SARS-CoV-2 in wastewater suggest a range
of $\mu_{ww}$ from 1e-8 to 1e-4.
This seems like a reasonable starting point, but we would like to include a lot more uncertainty.

We use:

$$\text{logit}(\mu_{ww}) \sim \text{Normal}(-14, 6)$$
With these parameters:

 -  The median is 8.3e-7. This corresponds to an RA(1%) of 8.3e-9,
    compared to our blog post MU Flu A estimates of 1.4e-8.
 -  70% of the probability is between 2e-9 and 3e-4.
 -  96% of the probability is between 5e-12 and 0.1,
    spanning more than 10 orders of magnitude.

Because these are respiratory viruses, we might expect the swabs to have higher relative abundance than wastewater at the same prevalence.
Accordingly, we'll shift the prior for $\mu_{swab}$ to the right relative to $\mu_{ww}$.
However, we don't want to make this assumption too strongly,
so we'll shift it by less than a standard deviation to:
$$\text{logit}(\mu_{swab}) \sim \text{Normal}(-10, 6)$$
This implies a ratio of the prior medians of $\approx 50$.

```{r}
#| label: prior-mu

# swab
mean_logit_mu_swab <- -10  # Prior mean on logit scale
sd_logit_mu_swab <- 6      # Prior standard deviation on logit scale
prior_logit_mu_swab <- function(x) dnorm(x, mean_logit_mu_swab, sd_logit_mu_swab)

# wastewater
mean_logit_mu_ww <- -14  # Prior mean on logit scale
sd_logit_mu_ww <- 6      # Prior standard deviation on logit scale
prior_logit_mu_ww <- function(x) dnorm(x, mean_logit_mu_ww, sd_logit_mu_ww)

crossing(
  datatype = factor(c("swab", "wastewater")),
  logit_mu = seq(-31, 5, length.out = 1000),
  ) %>%
  mutate(
    mu = plogis(logit_mu),
    prior = if_else(
      datatype == "swab",
      prior_logit_mu_swab(logit_mu),
      prior_logit_mu_ww(logit_mu)
    ),
  ) %>%
  ggplot(aes(mu, prior, color = datatype, linetype = datatype)) +
  geom_line() +
  scale_x_continuous(
    transform = "logit",
    breaks = c(10^seq(-12, -2, 2), 0.5), #, 0.99),
    name = expression(mu~"(logistic scale)"),
  ) +
  labs(title = expression("Priors on "~mu))
```

### Overdispersion

Our final parameters are the inverse-overdispersion parameters for the read counts ($\phi_1$ and $\phi_{ww}$).
A useful intuition for these parameters is that, as the read depth gets very large so that Poisson counting noise becomes unimportant relative to other sources of variance, the coefficient of variation in read counts goes to $\phi^{-1/2}$.

We have little prior information about this parameter except that it must be greater than zero, and we expect significant overdispersion in both sample sources.
Another complication is that we don't expect our data, especially the swab data, to give us much information about $\phi$ because you need multiple non-zero observations to estimate overdispersion.

For wastewater, we'll accept our ignorance and use a weakly informative prior:
$$\phi_{ww} \sim \text{Gamma}(1, 1) = \text{Exponential}(1)$$.

For swabs, the most important impact of $\phi_1$ on our estimates of prevalence
comes from the fact that it, along with $\mu_{swab}$, controls the probability of getting zero viral reads from a positive swab.
This relates to a major source of overdispersion: variation in how well subjects swab themselves.
From field observations, Simon estimates that roughly 20% of people take a bad sample.
We can use this to get a sense of the reasonable range of $\phi_{swab}$.

Consider a case where you have a single positive swab in the pool,
and where sampled deeply enough that you expect to see roughly 100 viral reads, conditional on getting any at all (as a proxy for a getting a good swab).
Let $\lambda$ be the expected number of viral reads from a positive swab.
Then if the probability of getting zero viral reads is dominated by a 20% chance of getting a bad swab,
$$\lambda = 0 \times P(r_{viral} = 0 | \lambda, \phi_1) + 100 \times P(r_{viral} > 0 | \lambda, \phi_1) = 80$$
In our negative binomial model,
$$P(0 | \lambda, \phi_1) = {\left(\frac{\phi_1}{\lambda + \phi_1}\right)}^{\phi_1} = 0.2$$

Solving for $\phi_1$, we get $\phi_1 \approx 0.3$.

Of course, the 20% probability of a bad swab was just a guess.
If $P(r_{viral} = 0)$ is instead 0.1 or 0.4, we can repeat this exercise and get
$\phi_1 \approx 0.4$ and $\phi_1 \approx 0.15$ respectively.

As an extreme case, if the probability that you get zero reads from a positive swab
goes to 90% when you'd expect 100 if you got any,
then $\phi_1 \approx 0.016$.

With these values in mind, we select a prior
$$\phi_1 \sim Gamma(2, 7)$$,
which looks like:

```{r}
#| label: prior-phi

# swab
shape_phi_1 <- 2
rate_phi_1 <- 7
prior_phi_1 <- function(x) dgamma(x, shape_phi_1, rate_phi_1)

# wastewater
shape_phi_ww <- 1
rate_phi_ww <- 1
prior_phi_ww <- function(x) dgamma(x, shape_phi_ww, rate_phi_ww)

crossing(
  datatype = factor(c("swab", "wastewater")),
  phi = seq(0.01, 10, 0.001)
) %>%
  mutate(
    prior = if_else(datatype == "swab", prior_phi_1(phi), prior_phi_ww(phi)),
    prior_log10 = prior * phi * log(10), # Jacobian adjustment
  ) %>%
  ggplot(aes(phi, prior_log10, color = datatype, linetype = datatype)) +
  geom_line() +
  geom_vline(xintercept = c(0.15, 0.28, 0.4), linetype = "dotted", color = "grey") +
  scale_x_log10() +
  labs(title = expression("Priors on"~phi), y = "", x = expression(phi))
```

Dotted grey lines show the values of $\phi$ that imply the probability of getting zero reads from a positive
swab is 0.4, 0.2, and 0.1 from left to right
(assuming that we expect to see 100 viral reads from a viral swab conditional on seeing any).

We can also calculate the implied distribution of the coefficient of variation of viral reads
in the deep sequencing limit:

```{r}
crossing(
  datatype = factor(c("swab", "wastewater")),
  cv = seq(0.1, 5, 0.01),
) %>%
  mutate(
    phi = cv^-2,
    prior = if_else(
      datatype == "swab",
      prior_phi_1(phi) * 2 * cv^-3, # Jacobian adjustment
      prior_phi_ww(phi) * 2 * cv^-3
      )
    ) %>%
  ggplot(aes(cv, prior, color = datatype, linetype = datatype)) +
  geom_line() 
```

In the future, we may want to consider fitting a model to all viruses together and estimating a single pair of $\phi$ parameters for all viruses (or a hierarchical model that partially pools information). This could make sense because many of the sources of noise, including swab quality, are probably independent of the virus.
However, the most straightforward such extension of the model may be computationally expensive,
so we would need to make an approximation or figure out a clever algorithm.

## Prior predictive checks

```{r}
#| label: prior-pred-check-swab
#| cache: true

n_prior_samples <- 1000
prior_samples <- tibble(
  rep = 1:n_prior_samples,
  p = plogis(rnorm(mean = mean_logit_p, sd = sd_logit_p, n = n_prior_samples)),
  mu_swab = plogis(rnorm(mean = mean_logit_mu_swab, sd = sd_logit_mu_swab, n = n_prior_samples)),
  phi_1 = rgamma(shape = shape_phi_1, rate = rate_phi_1, n = n_prior_samples),
) %>%
  cross_join(swab_metadata) %>%
  mutate(
    n_pos = map2_int(pool_size, p, ~ rbinom(1, .x, .y)),
    mu = n_pos * mu_swab * all_reads / pool_size,
    phi = n_pos * phi_1,
    viral_reads = map2_int(mu, phi, ~ ifelse(.x > 0, rnbinom(
      1, size = .y, mu = .x
    ), 0)),
  ) %>%
  glimpse

prior_predictive_summary <- prior_samples %>%
  summarise(
    pos_swabs = sum(n_pos),
    pos_pools = sum(viral_reads > 1),
    exp_pos_swabs = sum(p * pool_size),
    p_no_pos_swabs = exp(-exp_pos_swabs),
    total_reads = sum(viral_reads),
    max_reads = max(viral_reads),
    .by = rep,
  ) 

prior_predictive_summary %>%
  summarise(
    frac_no_pos_swabs = mean(pos_swabs == 0),
    frac_no_pos_pools = mean(pos_pools == 0),
  )

prior_predictive_summary %>%
  ggplot(aes(x = pos_pools)) +
  geom_histogram(binwidth = 1) +
  geom_point(aes(y = 0), data = species_swab_summary, position = position_jitter()) 

prior_predictive_summary %>%
  filter(pos_pools > 0) %>%
  ggplot(aes(x = total_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_total, y = 0), data = species_swab_summary) +
  scale_x_continuous(transform = "log10", breaks = 10^seq(0:5))

prior_predictive_summary %>%
  filter(pos_pools > 0) %>%
  ggplot(aes(x = max_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_max, y = 0), data = species_swab_summary) +
  scale_x_continuous(transform = "log10", breaks = 10^seq(0:4))
```

```{r}
#| label: prior-pred-check-ww
#| cache: true

prior_samples_ww <- tibble(
  rep = 1:n_prior_samples,
  p = plogis(rnorm(mean = mean_logit_p, sd = sd_logit_p, n = n_prior_samples)),
  mu_ww = plogis(rnorm(mean = mean_logit_mu_ww, sd = sd_logit_mu_ww, n = n_prior_samples)),
  phi_ww = rgamma(shape = shape_phi_ww, rate = rate_phi_ww, n = n_prior_samples),
) %>%
  cross_join(swab_metadata) %>%
  mutate(
    mu = p * mu_ww * all_reads,
    viral_reads = map2_int(mu, phi_ww, ~ rnbinom(
      1, size = .y, mu = .x
    )),
  ) %>%
  glimpse

prior_predictive_summary_ww <- prior_samples_ww %>%
  summarise(
    pos_samples = sum(viral_reads > 1),
    total_reads = sum(viral_reads),
    max_reads = max(viral_reads),
    .by = rep,
  ) 

prior_predictive_summary_ww %>%
  summarise(mean(pos_samples == 0))

prior_predictive_summary_ww %>%
  ggplot(aes(x = pos_samples)) +
  geom_histogram(binwidth = 1) +
  geom_point(aes(y = 0), data = species_ww_summary, position = position_jitter()) 

prior_predictive_summary_ww %>%
  filter(pos_samples > 0) %>%
  ggplot(aes(x = max_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_max, y = 0), data = species_ww_summary) +
  scale_x_continuous(transform = "log10")

prior_predictive_summary_ww %>%
  filter(pos_samples > 0) %>%
  ggplot(aes(x = total_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_total, y = 0), data = species_ww_summary, position = position_dodge()) +
  scale_x_continuous(transform = "log10")
```

## Stan model

Now we'll define the Stan model code.

```{r}
#| label: model
#| code-fold: false
#| cache: true

model_code = "
data {
  // Pooled swab sequence data
  int<lower=1> N_pools;                   // number of pools
  int<lower=1> n_swabs[N_pools];          // number of swabs in each pool
  int<lower=0> r_total_pool[N_pools];     // total read count
  int<lower=0> r_viral_pool[N_pools];     // viral read count
  
  // Wastewater MGS data
  int<lower=0> N_ww;                      // number of wastewater samples
  int<lower=0> r_total_ww[N_ww];          // total read count
  int<lower=0> r_viral_ww[N_ww];          // viral read count
}
  
parameters {
  real logit_p;       // prevalence
  
  // Swab parameters (using logit transformation)
  real logit_mu_swab;                // logit of expected viral fraction in a pool containing a single positive swab
  real<lower=0> phi_1;            // inverse-dispersion in a pool containing a single positive swab
  
  // Wastewater MGS parameters (using logit transformation)
  real logit_mu_ww;               // logit of expected viral fraction in wastewater if everyone were infected
  real<lower=0> phi_ww;           // inverse-dispersion in wastewater MGS
}

transformed parameters {
  real<lower=0, upper=1> p = inv_logit(logit_p);          // back-transform to [0,1] scale
  real<lower=0, upper=1> mu_swab = inv_logit(logit_mu_swab);    // back-transform to [0,1] scale
  real<lower=0, upper=1> mu_ww = inv_logit(logit_mu_ww);  // back-transform to [0,1] scale
}
  
model {
  // Priors
  logit_p ~ normal(${mean_logit_p}, ${sd_logit_p});
  logit_mu_swab ~ normal(${mean_logit_mu_swab}, ${sd_logit_mu_swab});
  phi_1 ~ gamma(${shape_phi_1}, ${rate_phi_1});
  logit_mu_ww ~ normal(${mean_logit_mu_ww}, ${sd_logit_mu_ww});
  phi_ww ~ gamma(${shape_phi_ww}, ${rate_phi_ww});  // Updated from placeholder
  
  // Marginalized likelihood for pooled testing data
  for (i in 1:N_pools) {
    vector[n_swabs[i] + 1] lpmf_sum;  // to store log probabilities for each possible n_pos value
    for (n_pos in 0:n_swabs[i]) {
      // Probability of n_pos positive samples out of n_swabs[i]
      real binomial_lpmf_val = binomial_lpmf(n_pos | n_swabs[i], p);
      
      // Probability of observing r_viral_pool[i] given n_pos positive samples
      real nb_lpmf_val; 
      if (n_pos == 0){
        // If no positive samples in the pool, we should see zero viral reads
        nb_lpmf_val = r_viral_pool[i] == 0 ? 0 : negative_infinity(); 
      } else {
        nb_lpmf_val = neg_binomial_2_lpmf(r_viral_pool[i] | r_total_pool[i] * n_pos * mu_swab / n_swabs[i], n_pos * phi_1);
      }
      
      // Store the sum of log probabilities
      lpmf_sum[n_pos + 1] = binomial_lpmf_val + nb_lpmf_val;
    }
    
    // Marginalize out n_pos[i] using log-sum-exp for numerical stability
    target += log_sum_exp(lpmf_sum);
  }
  
  // Wastewater MGS
  for (j in 1:N_ww) {
    real mu = r_total_ww[j] * p * mu_ww;
    r_viral_ww[j] ~ neg_binomial_2(mu, phi_ww);
  }
}
"
stan_model <- stan_model(model_code = str_interp(model_code))
```

# Fitting the model
## Model Fitting

Now we'll assemble the complete dataset and fit the model to each virus species:

```{r}
#| label: fit-model
#| cache: true
#| code-fold: false
#| echo: false

species_data <- full_join(
    swab_reads_complete %>% nest(.by = c(species, group)),
    wastewater_reads_complete %>% nest(.by = c(species, group)),
    by = join_by(species, group),
    suffix = c(".swab", ".wastewater")
)

create_stan_data <- function(df.sw, df.ww) {
  list(
    N_pools = nrow(df.sw),
    n_swabs = df.sw$pool_size,
    r_total_pool = df.sw$all_reads,
    r_viral_pool = df.sw$viral_reads,
    N_ww = nrow(df.ww),
    r_total_ww = df.ww$all_reads,
    r_viral_ww = df.ww$viral_reads
  )
}

fit_model <- function(d) {
  sampling(stan_model, d, iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 381928)
}

fits <- species_data %>%
  mutate(
    stan_data = map2(data.swab, data.wastewater, create_stan_data),
    fit = map(stan_data, fit_model)
  )
```

For model fitting, we use 4 chains run for 4,000 iterations each and a warmup of 1,000 iterations.

## Model Diagnostics

After fitting the model, it's important to check diagnostics to ensure the model is reliable.
These diagnostics help us assess the quality of the model fits:

- Divergent transitions indicate potential issues with the model
- Maximum treedepth indicates how complex the posterior exploration was
- Rhat values close to 1 indicate good convergence

```{r}
#| code-fold: false

get_diagnostics <- function(f) {
  check <- summary(f)$summary
  sampler_params <- get_sampler_params(f, inc_warmup = FALSE)
  
  n_divergent <- sum(do.call(rbind, sampler_params)[, "divergent__"])
  max_treedepth <- max(do.call(rbind, sampler_params)[, "treedepth__"])
  # Stan default max_treedepth = 10
  n_max_treedepth <- sum(do.call(rbind, sampler_params)[, "treedepth__"] == 10)
  rhat_max <- max(check[, "Rhat"], na.rm = TRUE)
  
  list(
    n_divergent = n_divergent,
    max_treedepth = max_treedepth,
    n_max_treedepth = n_max_treedepth,
    rhat_max = rhat_max
  )
}

fits %>%
  mutate(diagnostics = map(fit, get_diagnostics)) %>%
  select(species, group, diagnostics) %>%
  unnest_wider(diagnostics) %>%
  kable
```

No problems here.

We extract the posterior samples for further analysis, and write to text for later:

```{r}
#| code-fold: false
posteriors <- fits %>%
  mutate(posterior = map(fit, compose(as_tibble, extract))) %>%
  select(species, group, posterior) %>%
  unnest(cols = posterior) %>%
  mutate(
    across(-c(species, group), as.double),
    )

write_tsv(posteriors, "posteriors.tsv")
```

# Posterior distributions

## Marginal posterior distributions

Let's visualize the posterior distributions for our key parameters.
In the following plots, solid lines show the posterior distribution of prevalence for each virus species, compared to the prior distribution (dashed line).
Colors represent viral groups (SARS-CoV-2, seasonal Coronaviruses, Influenza, Mononegavirales, and Rhinoviruses).

For reference, here's a summary of the data:

```{r}
full_join(
  swab_reads_complete %>%
    summarise(
      `+ pools` = sum(viral_reads > 0),
      `swab reads` = sum(viral_reads),
      .by = species,
    ),
  wastewater_reads_complete %>%
    summarise(
      `+ ww samples` = sum(viral_reads > 0),
      `ww reads` = sum(viral_reads),
      .by = species,
    ),
  by = join_by(species)
  ) %>%
  kable
```

### Prevalence

```{r}
# Helper function for plotting priors
prior_density <- function(x, density_function, log10x = FALSE){
  if (log10x){
    data <- tibble(
      x = x,
      density = density_function(x) * x * log(10), # Jacobian adjustment
    )
  } else {
    data <- tibble(
      x = x,
      density = density_function(x),
    )
  }
  geom_line(
    aes(x, density),
    data = data,
    color="grey",
    linetype="dashed",
    inherit.aes = FALSE
    )
}
```

```{r}
posteriors %>%
  ggplot(aes(p, color=group)) +
  geom_density() +
  prior_density(plogis(seq(-12, 4, length.out=100)), compose(prior_logit_p, qlogis)) +
  scale_x_continuous(transform = "logit", limits = c(1e-5, 2e-1), breaks = 10^seq(-4, -1, 1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  labs(x = "prevalence", y = "") +
  theme(legend.position = "", axis.text.y = element_blank()) 
```

We can also plot the log density:

```{r}
posteriors %>%
  ggplot(aes(p, color=group)) +
  geom_density() +
  prior_density(plogis(seq(-12, 4, length.out=100)), compose(prior_logit_p, qlogis)) +
  scale_x_continuous(transform = "logit", limits = c(1e-5, 2e-1), breaks = 10^seq(-4, -1, 1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(
    transform = "log10",
    limits = c(1e-2, 1),
    breaks = 10^seq(0, -2),
    labels = scientific
  ) +
  labs(x = "prevalence", y = "") +
  theme(legend.position = "")
```

Comparison with perfect sensitivity maximum likelihood estimate: 

```{r}
perfect_sensitivity <- swab_reads_complete %>%
    summarise(
      pos_pools = sum(viral_reads > 0),
      p = pos_pools / (sum(pool_size[viral_reads == 0]) + sum(pool_size[viral_reads >0]) / 2),
      sd_logp = 1 / sqrt(pos_pools),
      viral_reads = sum(viral_reads),
      .by = c(species, group),
      )

posteriors %>%
  summarise(median(p), sd(log(p)), .by = c(species, group)) %>%
  left_join(
    perfect_sensitivity,
  ) %>%
  mutate(
    p_ratio = `median(p)` / p,
    sd_ratio = `sd(log(p))` / sd_logp,
    .keep = "unused",
  ) %>%
  kable(digits = 1)
```

```{r}
posteriors %>%
  ggplot(aes(x = p, y = species, fill = group)) +
  geom_vline(xintercept = plogis(mean_logit_p), color = "grey") +
  geom_vline(xintercept = plogis(mean_logit_p + sd_logit_p * c(1, -1)), color = "grey", linetype = "dashed") +
  geom_violin(draw_quantiles = c(0.15, .5, 0.85)) + 
  geom_point(aes(x = p), data = perfect_sensitivity, color = "white") +
  geom_errorbarh(aes(xmin = p / exp(sd_logp), xmax = p * exp(sd_logp)), data = perfect_sensitivity, color = "white") +
  scale_x_log10() +
  scale_y_discrete(limits=rev(species_order))
```

### Expected viral fractions

Posteriors on $\mu_{swab}$ (swab viral fraction with all positive swabs).

With linear y-scale:

```{r}
breaks <- 10^seq(-5, -1, 2)
posteriors %>%
  ggplot(aes(logit_mu_swab, color = group)) +
  geom_density() +
  prior_density(seq(-16, 2, length.out = 100), prior_logit_mu_swab) +
  scale_x_continuous(
    limits = c(-14, 2),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "", x = "mu_swab") +
  theme(legend.position = "", axis.text.y = element_blank()) 
```

And log y-scale:

```{r}
posteriors %>%
  ggplot(aes(logit_mu_swab, color = group)) +
  geom_density() +
  prior_density(seq(-16, 2, length.out = 100), prior_logit_mu_swab) +
  scale_x_continuous(
    limits = c(-14, 2),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "mu_swab") +
  theme(legend.position = "")
```

For species with multiple positive pools, (e.g. the Rhinoviruses), there is information to constrain the expected viral fraction in a positive swab.
With only one positive pool (e.g. H1N1, which has only one viral read), there is less, but still some information.

Note that all of the posteriors are shifted to the right relative to the prior.
This means that the prior is biasing these estimates downward.
This is a predictable effect of using the same prior for $\mu_{swab}$ and $\mu{ww}$
We may want to consider having separate priors.

HCoV-229E is an outlier because it has only a single positive pool with 6008 viral reads.
This puts most of the weight on $\mu_{swab} > 0.1$, which would be higher if the prior weren't pulling it down.
I'm not sure whether this is a good effect: regularization, or a bias we'd like to eliminate.
It depends on whether we think it's likely that $\mu_{swab}$ could actually be this high, vs this being a very lucky sample.

Same plots but for $\mu_{ww}$:

```{r}
breaks <- 10^seq(-7, -3, 2)
posteriors %>%
  ggplot(aes(logit_mu_ww, color = group)) +
  geom_density() +
  prior_density(seq(-20, -5, length.out = 100), prior_logit_mu_ww) +
  scale_x_continuous(
    limits = c(-20, -5),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  scale_y_continuous(name = "") +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "", x = "mu_ww") +
  theme(legend.position = "", axis.text.y = element_blank()) 
```

```{r}
posteriors %>%
  ggplot(aes(logit_mu_ww, color = group)) +
  geom_density() +
  prior_density(seq(-20, -5, length.out = 100), prior_logit_mu_ww) +
  scale_x_continuous(
    limits = c(-20, -5),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "mu_ww") +
  theme(legend.position = "")
```

Here the posteriors are narrow compared to the prior and not far from the prior mode.

### Overdispersion

Posteriors for swab inverse-overdispersion $\phi_1$.

```{r}
posteriors %>%
  ggplot(aes(phi_1, color=group)) +
  geom_density() +
  prior_density(10^seq(-3, 3, length.out=100), prior_phi_1, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1)) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "", axis.text.y = element_blank())
```

The posterior distributions are all very close to the prior,
suggesting that, as expected the data contains very little information about $\phi_1$, at least relative to the prior.

There is a slight leftward shift for HCoV-OC43, which has the most positive pools (5).

Posteriors for wastewater inverse-overdispersion $\phi_{ww}$.

```{r}
posteriors %>%
  ggplot(aes(phi_ww, color=group)) +
  geom_density() +
  prior_density(10^seq(-2, 3, length.out=100), prior_phi_ww, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "", axis.text.y = element_blank())
```

The wastewater data has more non-zero counts and thus more data to infer dispersion.
I'm not sure I trust any of these though.

Generally, the species that have high predicted overdispersion relative to the prior (low $\phi$) are ones with fewer positive wastewater samples. 
That could be because there are more zero-read samples than expected given the counts in the non-zero-read samples.
Or it could be because the estimates are noisier with less data.

## Joint Posterior Distributions

We can also examine relationships between different parameters in the posterior.
In the following plots, each point is a sample from the joint posterior of two parameters.
The curves are an estimate of the conditional mean of the y-axis parameter given the x-axis parameter.

### Realtionships with prevalence

Prevalence vs $\mu_{swab}$ (note log x-scale not logit):

```{r}
posteriors %>%
  ggplot(aes(mu_swab, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(limits = c(1e-7, 1), labels = scientific) +
  scale_y_log10(limits = c(1e-5, 0.5), labels = scientific) +
  facet_wrap(~species)
```

It looks like:

- For small $\mu$, prevalence grows as $\mu$ decreases.
- For larger $\mu$, this relationship flattens.

This makes sense because low $\mu$ requires high prevalence to observe the virus since each positive swab is likely to go unobserved, while with high $\mu$ positive swabs are usually observed, so the prevalence is determined by the number of positive pools.

TODO: Add maximum likelihood calculation here.

```{r}
posteriors %>%
  ggplot(aes(phi_1, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(limits = c(1e-3, 10), labels=scientific) +
  scale_y_log10(limits = c(1e-5, 0.5), labels = scientific) +
  facet_wrap(~species)
```

Smaller $\phi_1$ implies a higher probability of failing to get viral reads a positive swab.
Thus, the inferred prevalence conditional on $phi_1$ is inversely related to $phi_1$.

```{r}
posteriors %>%
  summarise(
    corr_log_p_phi_1 = cor(log(p), log(phi_1), method = "spearman"),
    corr_p_phi_1 = cor(p, phi_1, method = "spearman"),
    corr_p_mu_ww = cor(log(p), log(mu_swab), method = "spearman"),
    .by = species,
  )
```

To encapsulate the relationship between $p$, $\mu_{swab}$, and $\phi_1$,
we can approximate $\hat{p}$, the maximum likelihood $p$ conditional on known $\mu_{swab}$ and
$\phi_1$.
(The approximation is in the the limit where most pools are expected to be negative.)

```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_1 / (lambda + phi_1)) ^ phi_1,
  ) %>%
  summarise(
    p = first(p),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  ggplot(aes(p_hat, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() + 
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  scale_x_log10(limits = c(2e-3, 1)) +
  scale_y_log10() +
  facet_wrap(~species)
```
```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_1 / (lambda + phi_1)) ^ phi_1,
  ) %>%
  summarise(
    p = first(p),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  summarise(
    corr_sp = cor(log(mean_p0), log(p), method = "spearman"),
    corr_pe = cor(log(mean_p0), log(p), method = "pearson"),
    .by = c(species),
  )
```

The approximate maximum likelihood and posterior $p$ qualitatively agree.

### Relationships with expected wastewater viral fraction

Prevalence: 

```{r}
posteriors %>%
  ggplot(aes(p, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```

```{r}
posteriors %>%
  summarise(
    cor_sp = cor(p, mu_ww, method = "spearman"),
    cor_log = cor(log(p), log(mu_ww), method = "pearson"),
    .by = species,
  )
```

As expected, there is a very tight relationship between prevalence and estimated wastewater viral fraction (per prevalence).
Often, strong correlations between parameters suggest that we should reparameterize the model so the parameters are more independent.
However, in this case, I think this is appropriate. We need a distinct $p$ parameter to make use of the discreteness of the swab pools.

The next three plots show the relationship between $\mu_{ww}$ and $\mu_{swab}$, $\phi_1$, and $\phi_{ww}$:

```{r}
posteriors %>%
  ggplot(aes(mu_swab, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```


```{r}
posteriors %>%
  ggplot(aes(phi_1, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```

```{r}
posteriors %>%
  ggplot(aes(phi_ww, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10() + # label = scientific) +
  scale_y_log10(label = scientific) +
  facet_wrap(~species)
```

The relationships with $\mu_{swab}$ and $\phi_1$ are mediated by $p$ and are not surprising.

The relationship with $\phi_{ww}$ suggests that the posteriors are uncorrelated,
which is encouraging because we don't have any real information about this parameter.

# Posterior Predictive Checks

## Swab data

To validate our model, we can use posterior predictive checks - generating new data from the fitted model and comparing it to the observed data.
The following table gives the summary statistics for the data.

```{r}
#| label: swab-ppd
#| cache: true

posterior_predictive_swab <- posteriors %>%
  mutate(rep = 1:n(), .by = species) %>%
  cross_join(swab_metadata) %>%
  mutate(
    n_pos = map2_int(pool_size, p, ~ rbinom(1, .x, .y)),
    mu = n_pos * mu_swab * all_reads / pool_size,
    phi = n_pos * phi_1,
    viral_reads = map2_int(mu, phi, ~ ifelse(.x > 0, rnbinom(
      1, size = .y, mu = .x
    ), 0)),
    false_negative = n_pos > 0 & viral_reads == 0,
  )

posterior_predictive_swab_summary <- posterior_predictive_swab %>%
  summarise(
    pos_pools = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    false_negative_pools = sum(false_negative),
    across(c(p, mu_swab, phi_1), first),
    .by = c(species, group, rep),
  )



kable(species_swab_summary)
```


```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = pos_pools, fill = group, after_stat(density))) +
  geom_histogram(binwidth = 1, center = 0, color = "grey") +
  geom_vline(data = species_swab_summary, mapping = aes(xintercept=pos_pools)) +
  facet_wrap(vars(species)) +
  scale_x_continuous(breaks = seq(0, 12, 2), name = "positive pools")
```

This plot compares the observed number of positive pools (vertical line) with the distribution of positive pools generated from our model.
For every virus, the observed number of positive pools is common in the simulations.

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_total, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_continuous(
    transform = "log10",
    breaks = 10^seq(0, 4),
    name = "total viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed total viral reads (points) with the distribution from our model.
Vertical bars show quartiles

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_max, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_continuous(
    # transform = "pseudo_log",
    # breaks = c(0, 10^seq(0, 4)),
    transform = "log10",
    breaks = 10^seq(0, 4),
    name = "largest viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed maximum viral reads with the distribution from our model.

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_sd, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_log10(
    breaks = 10^seq(0, 4),
    name = "stdev viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed standard deviation of viral reads with the distribution from our model.
Violins show only the simulations with at least one viral read.
Vertical bars show quartiles (conditional on at least one read).

Together, these plots suggest that the model is capturing the swab data reasonably well.

## Wastewater data

Now we'll perform similar posterior predictive checks for the wastewater data:

```{r}
#| label: wastewater-ppd
#| cache: true

posterior_predictive_ww <- posteriors %>%
  mutate(rep = 1:n(), .by=species) %>%
  cross_join(wastewater_metadata) %>%
  mutate(
    mu = p * mu_ww * all_reads,
    viral_reads = map2_int(mu, phi_ww, ~ ifelse(.x > 0, rnbinom(1, size = .y, mu = .x), 0)), 
  )

posterior_predictive_ww_summary <- posterior_predictive_ww %>%
  summarise(
    pos_samples = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    across(c(p, mu_ww, phi_ww), first),
    viral_reads_expected = p * mu_ww * sum(all_reads),
    .by = c(species, group, rep),
  )


kable(species_ww_summary)
```

This code calculates summary statistics for the observed wastewater data, which we'll use to compare with predictions from our model.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = pos_samples, fill = group, after_stat(density))) +
  geom_histogram(binwidth = 1, center = 0, color = "grey") +
  geom_vline(data = species_ww_summary, mapping = aes(xintercept = pos_samples)) +
  facet_wrap(vars(species)) +
  scale_x_continuous(breaks = seq(0, 12, 2), name = "positive samples")
```

This plot compares the observed number of positive wastewater samples (vertical line) with the distribution generated from our model.
It looks like the model systematically overestimates the number of wastewater samples with viral reads (though not by a lot in any one case).
This may be a sign that $\phi_{ww}$ is too large: too little overdispersion leading to too few zero counts given a mean relative abundance.
It could also mean that the negative binomial model is slightly misspecified.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_total, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed total viral reads in wastewater (points) with the distribution predicted by our model.
The observations here are all very close to the posterior median.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_max, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed maximum viral reads in wastewater with the distribution from our model.

Here the model might be systematically slightly overestimating the maximum viral reads.
This cuts against the "too little overdispersion" hypothesis for the excess positive samples.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_sd, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed standard deviation of viral reads in wastewater with the distribution from our model.

Here the model might also be systematically slightly too high.

Taking these together, it could be that the negative binomial model can't simultaneously fit the number
of positive samples and the dispersion of reads because these are controlled by a single parameter.

I don't think this is a major issue for trusting the results.
The observed values are all well within the posterior predictive distributions
and it probably makes sense to be conservative with respect to the possibility of seeing anomalously large counts. 

# Results 

## Prevalence

```{r}
posteriors %>%
  ggplot(aes(p, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(name = "prevalence", limits = c(1e-4 / 4, 4e-1)) +
  scale_y_discrete(limits=rev(species_order))
```

```{r}
prev_posteriors <- left_join(
  swab_reads_complete %>%
    summarise(
      `+ pools` = sum(viral_reads > 0),
      `viral reads` = sum(viral_reads),
      .by = species,
    ),
  posteriors %>%
    summarise(
      q15 = quantile(p, .15),
      median = median(p),
      q85 = quantile(p, .85),
      `q85 / q15` = q85 / q15,
      .by = species,
    ),
  )

write_tsv(prev_posteriors, "prevalence_posteriors.tsv")

prev_posteriors %>%
  kable(digits = c(NA, 0, 0, 4, 4, 4, 2)) %>%
  column_spec(2:7, monospace = TRUE)
```
```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_1 / (lambda + phi_1)) ^ phi_1,
  ) %>%
  summarise(
    p = first(p),
    mu_swab = first(mu_swab),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  summarise(
    median_mu_swab = median(mu_swab),
    median_p0 = median(mean_p0),
    .by = species,
  ) 
```

Let's compare these results to what we'd get with perfect sensitivity.
In that case, we can estimate the maximum likelihood $p$ and the Fisher information for $\log p$,
which gives an estimate of the posterior posterior mode and inverse variance of $\log p$ if we had a flat prior on $\log p$.

### Comparison with perfect sensitivity maximum likelihood estimate: 

```{r}
perfect_sensitivity <- swab_reads_complete %>%
    summarise(
      pos_pools = sum(viral_reads > 0),
      p = pos_pools / (sum(pool_size[viral_reads == 0]) + sum(pool_size[viral_reads >0]) / 2),
      sd_logp = 1 / sqrt(pos_pools),
      viral_reads = sum(viral_reads),
      .by = c(species, group),
      )

comparison <- posteriors %>%
  summarise(median(p), sd(log(p)), .by = c(species, group)) %>%
  left_join(
    perfect_sensitivity,
  ) %>%
  mutate(
    p_ratio = `median(p)` / p,
    sd_ratio = `sd(log(p))` / sd_logp,
    .keep = "unused",
  )

comparison %>%
  kable(digits = 1)

comparison %>%
  filter(group %in% c("Coronaviruses (SARS-CoV-2)", "Coronaviruses (seasonal)", "Rhinoviruses")) %>%
  summarise(
    n(),
    median(p_ratio),
    median(sd_ratio),
  )
```

```{r}
posteriors %>%
  ggplot(aes(x = p, y = species, fill = group)) +
  geom_vline(xintercept = plogis(mean_logit_p), color = "grey") +
  geom_vline(xintercept = plogis(mean_logit_p + sd_logit_p * c(1, -1)), color = "grey", linetype = "dashed") +
  geom_violin(draw_quantiles = c(0.15, .5, 0.85)) + 
  geom_point(aes(x = p), data = perfect_sensitivity, color = "white") +
  geom_errorbarh(aes(xmin = p / exp(sd_logp), xmax = p * exp(sd_logp)), data = perfect_sensitivity, color = "white") +
  scale_x_log10() +
  scale_y_discrete(limits=rev(species_order))
```

Here, the violins show posteriors with median, 15-th, and 85-th percentiles as black lines.
The white dots are the maximum likelihood estimates with error bars $\log \hat{p} \pm {\mathcal{I_{\log p}}}^{-1/2}$.
Solid grey line is the prior median and dashed grey lines are the prior 15th and 85th percentiles.

## RA_p(1%)

Finally, we can calculate the RA_p(1%) values - the relative abundance at 1% prevalence - which was a key goal of the original analysis:

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  ggplot(aes(ra01, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(name = "wastewater RA_p(1%)") +
  scale_y_discrete(limits=rev(species_order))
```

This plot shows the posterior distribution of RA_p(1%) for each virus species.
Vertical lines represent 15%, 50%, and 85% quantiles of the posterior.

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q05 = quantile(ra01, 0.05),
    q15 = quantile(ra01, 0.15),
    q25 = quantile(ra01, 0.25),
    median = median(ra01),
    q75 = quantile(ra01, 0.75),
    q85 = quantile(ra01, 0.85),
    q95 = quantile(ra01, 0.95),
    ratio_q85q15 = q85 / q15,
    gsd_sq = exp(2 * sd(log(ra01))),
    ratio_q95q05 = q95 / q05,
    .by = c(species, group),
  ) %>%
  mutate(
    across(
      -c(species, group, starts_with("ratio_"), gsd_sq),
      ~ format(.x, scientific = TRUE, digits = 2)
    ),
    across(
      c(starts_with("ratio_"), gsd_sq),
      ~ format(.x, digits = 2)
    )
  ) %>% 
  select(-group) %>%
  kable
```

This table provides quantiles of the RA_p(1%) estimates for each virus species, representing the uncertainty in these values.

```{r}
ra_posteriors <- posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q15 = quantile(ra01, .15),
    median = median(ra01),
    q85 = quantile(ra01, .85),
    `q85 / q15` = q85 / q15,
    .by = c(species)
  )

write_tsv(ra_posteriors, "ra_posteriors.tsv")

ra_posteriors %>% 
  mutate(
    across(
      c(q15, median, q85),
      ~ format(.x, scientific = TRUE, digits = 2)
    ),
    across(
      c(`q85 / q15`),
      ~ format(.x, digits = 2)
    )
  ) %>%
  kable(align = "lrrrr") %>%
  column_spec(2:5, monospace = TRUE)
```

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q15 = quantile(ra01, .15),
    median = median(ra01),
    q85 = quantile(ra01, .85),
    sd = exp(sd(log(ra01))),
    `q85 / q15` = q85 / q15,
    .by = c(species)
  ) %>%
  summarise(
    exp(sd(log(median))),
    median(sd),
  )
```


## RA swab / RA wastewater

```{r}
posteriors %>%
  mutate(ratio = mu_swab / mu_ww) %>%
  ggplot(aes(ratio, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(
    name = "swab RA / wastewater RA",
    limits = c(1e-2, 1e8),
    breaks = 10^seq(0, 6, 2)
  ) +
  scale_y_discrete(limits=rev(species_order))
```

This plot shows the ratio of swab relative abundance to wastewater relative abundance for a fixed prevalence.
Vertical lines represent quartiles of the posterior.

As a sanity check, let's calculate some two different estimates of this factor from the count data alone:
1. The ratio of relative abundances, summing reads over samples
2. The ratio of relative abundances, averaging RA over samples

```{r}
get_ra <- function(data){
  summarise(
    data,
    ra_by_read = sum(viral_reads) / sum(all_reads),
    ra_by_sample = mean(viral_reads / all_reads),
    .by = c(species, group)
  )
}

naive_ra_ratios <- left_join(
  get_ra(swab_reads_complete),
  get_ra(wastewater_reads_complete),
  by = join_by(species, group),
  suffix = c(".swab", ".ww"),
) %>%
  mutate(
    ratio_by_read = ra_by_read.swab / ra_by_read.ww,
    ratio_by_sample = ra_by_sample.swab / ra_by_sample.ww,
  )

posteriors %>%
  mutate(ratio = mu_swab / mu_ww) %>%
  ggplot(aes(ratio, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  geom_point(aes(ratio_by_read, species), naive_ra_ratios, shape = "x") +
  geom_point(aes(ratio_by_sample, species), naive_ra_ratios, shape = "o") +
  scale_x_log10(
    name = "swab RA / wastewater RA",
    limits = c(1e-2, 1e8),
    breaks = 10^seq(0, 6, 2)
  ) +
  scale_y_discrete(limits=rev(species_order))
```

I don't see a systematic bias here, which suggests that the priors aren't influencing the ratio too much.
