---
title: "Theoretical analysis of the precision to estimate prevalence"
format:
  html:
    code-fold: true
    toc: true
execute: 
  cache: true
editor: source
author: dan
date: 2025-05-15
---

An important question about our model is: why does our model estimate
the viral prevalence with finite precision given the fact that a pool
can have no viral reads because of low prevalence or low expected relative abundance?

We will try to establish that this precision comes from the combination of two things:

1. The likelihood
2. The prior on $\phi_1$, the inverse overdispersion for viral reads from a single positive swab,
through its control of the probability of getting zero viral reads from a positive swab.

Importantly, the precision does not depend on having a narrow prior on $p$ or
on $\mu_{swab}$, the expected relative abundunce in a positive swab.

# Analysis plan

To show that the precision comes from the likelihood, we will examine the likelihood function
alone without looking at a prior.
We will calculate the maximum likelihood estimate (MLE) and the Hessian (matrix of second derivatives)
of the log likelihood evaluated at the MLE,
which gives an estimate of the precision.
To lowest order, the likelihood function is approximately a Gaussian function with its maximum at the MLE
and its precision matrix (inverse covariance) given by the Hessian.

To simplify our calculations, we will assume $\phi_1$ known.
Our posterior samples show that the data we have does not constrain $\phi_1$.
We can then vary the value of $\phi_1$ and study how it changes the MLE and the precision.

# The likelihood

Consider a set of pools indexed by $i$.
In each pool, we observe $r_i \in Z^+$ viral reads.
For simplicity, assume that each pool contains $s$ swabs and that each is sequenced to the same depth $d$.
To avoid carrying around constants,
let $\lambda = \frac{d}{s} \mu_{swab}$ be the expected number of viral reads in a pool if one swab is positive.
Estimating $\lambda$ is equivalent to estimating $\mu_{swab}$.

Let $p$ be the prevalence.
Let $\phi_1$ be the inverse-overdispersion of viral read counts from a single infected swab.

Assuming the pools are independent, the log likelihood is a sum over pools:

$$\log L(p, \lambda, \phi_1 | r) = \sum_i \log Pr(r_i | p, \lambda, \phi_1)$$ {#eq-likelihood}

It is convenient to marginalize over the number of swabs
contributing viral reads to the pool, which we'll call $S_{>0}$:

$$P(r_i | p, \lambda, \phi_1) = \mathbb{E}[Pr(r_i | S_{>0}) | p, \lambda, \phi_1]$$ {#eq-marginalize}

(Note that $S_{>0}$ is less than or equal to the number of swabs from infected people.)
Assuming that swabs are independent,
$$S_{>0} \sim \text{Binomial}\left(s, p (1 - P(0 | \lambda, \phi_1))\right),$$
where $P(0 | \lambda, \phi_1)$ is the probability that a single swab from an infected person contributes zero reads.
(A note on notation: throughout, we use $Pr$ to represent a generic probability and $P$ to be the
probability mass function of read counts from a single positive swab.)

From our data, we observe that most of the pools have zero reads.
This implies that 
$$\mathbb{E}(S_{>0}) = s p (1 - P(0 | \lambda, \phi_1)) < 1.$$

Thus, we can simplify our calculation with the approxmation that
$$S_{>0} \sim \text{Bernoulli}(s p (1 - P(0|\lambda, \phi_1))).$$

Applying this to @eq-marginalize, gives
$$P(r_i) = \begin{cases}
1 - s p (1 - P(0 | \lambda, \phi_1))\;\text{if}\;r_i = 0,\\
s p P(r_i | \lambda, \phi_i)\;\text{if}\;r_i > 0.
\end{cases}$$

Substituting into @eq-likelihood, and denoting the number of pools with zero and non-zero
viral reads as $n_0$ and $n_{>0}$ respectively, we get
$$\begin{align}
\log L & = n_0 \log(1 - sp(1 - P(0|\lambda, \phi_1))) + \sum_{\{i: r_i > 0\}} \log(sp P(r_i|\lambda, \phi_1))\\
       & \approx - n_0 s p (1 - P(0|\lambda, \phi_1)) + n_{>0} \log p + \sum_{\{i: r_i > 0\}} \log P(r_i|\lambda, \phi_1) + \text{const.}\\
\end{align}$$
where in the approximation we've kept only terms of order $sp(1-P(0))$ to be consistent with our Bernoulli approximation.

In our negative binomial model,
$$P(r) = \binom{r + \phi_1 - 1}{r} {\left(\frac{\lambda}{\lambda + \phi_1}\right)}^r {\left(\frac{\phi_1}{\lambda + \phi_1}\right)}^{\phi_1}$$ {#eq-nb}
so that
$$P(0 | \lambda, \phi_1) = {\left(\frac{\phi_1}{\lambda + \phi_1}\right)}^{\phi_1}.$$ {#eq-p0}

# Choosing values of $\phi_1$

We have to choose some values of $\phi_1$ to investigate.
We don't have much prior information about overdispersion in swab reads,
and fitting our model the data is not very informative either.
(Presumably this is because you need a lot of observation of counts to estimate overdispersion,
and we typically have only one or a few pools with viral reads.)

As we will see below, the primary way that $\phi_1$ influences the inference of $p$
is through $P(r = 0 | \lambda, \phi_1)$, the probability that a swab from an infected person
contributes zero reads.
So it makes sense to consider a range $\phi_1$ that span different values of $P(0)$.

One effect we would like to capture is that we expect to sometimes get
zero viral reads *even when we sequence deeply enough that we expect to observe the virus*.
For example, if an infected participant swabs badly, they may contribute no viral reads to the pool
even at very high sequencing depth.
We don't know how often this happens, but Simon has observed that roughly 20%
of participants appear to swab badly.
The rate could be higher or lower, but it's probably not close to 100% or swab sampling
would be useless.
Thus, we would like to allow for a wide range of values of $P(r = 0)$ on a linear scale between 0 and 1.

To calculate $P(r = 0 | \lambda, \phi_1)$, we also have to specify a $\lambda$.
We'll define "deep sequencing" to be depth such that $E(r | r > 0) = 100$
and for each $\phi_1$ find the value of $\lambda$ that meets this condition.
This is deep enough that without significant overdispersion, we would never
expect to get zero counts.
It's also the same scale of counts as some of our more abundant viruses, including SARS-CoV-2.
(Also note that for small $\phi_1$, i.e. large overdispersion, $P(0)$ is not very sensitive to $\lambda$ so the exact choice doesn't matter much.)

Using @eq-p0, and
$$\mathbb{E}[r] = \lambda = \mathbb{E}[r|r>0](1 - P(0))$$
for we can solve for $\lambda$ and $P(0)$
for a given $\phi_1$ and $\mathbb{E}[r | r>0]$:

```{r}
#| message: false

library(tidyverse)
library(knitr)
p0 <- function(lam, phi) (phi / (lam + phi))^phi

find_lam <- function(phi, rv, npos){
  fn <- \(lam) lam - (rv / npos) * (1 - p0(lam, phi))
  uniroot(fn, c(1e-4, 10000))$root
}

shape_phi_1 <- 2
rate_phi_1 <- 7
prior_phi_1 <- function(x) pgamma(x, shape_phi_1, rate_phi_1)

df_phi <- crossing(
  phi = seq(0.005, 1, 0.005),
  r = c(2, 100, 500),
) %>%
  mutate(
    lam = map2_dbl(phi, r, \(x, y) find_lam(x, y, 1)),
    p0 = p0(lam, phi),
    prior = prior_phi_1(phi),
  ) 
  
df_phi %>% 
  ggplot(aes(phi, p0, color = as.factor(r))) + 
  geom_line() +
  labs(y = "", x = "phi_1", title = "P(r = 0)", color = "E[r | r > 0]")
```

As we vary $\phi_1$ between zero and one, $P(r = 0)$ when the expected non-zero read count is in the hundreds also varies between zero and one.
(We show both $\mathbb{E}[r | r > 0] = 100$ and $500$ to demonstrate that the exact choice of large value doesn't matter much.)

Note that with shallow sequencing (relative to shedding, $\mathbb{E}[r | r > 0] = 2$),
$P(r=0)$ is much higher than with deep sequencing.
The correspondence between $P(r=0)$ for deep vs shallow sequencing is a consequence
of the negative binomial model, rather than a tunable parameter.

When fitting the model, we put a $\text{Gamma}(2, 7)$ prior on $\phi_1$,
whose CDF looks like:

```{r}
df_phi %>%
  ggplot(aes(phi, prior)) +
  geom_line() +
  labs(y = "", x = "phi_1", title = "Prior CDF on phi_1")
```

We can use the analysis above to convert this into an implied prior CDF on $P(r = 0)$:

```{r}
df_phi %>%
  ggplot(aes(p0, 1 - prior, color = as.factor(r))) +
  geom_line() +
  labs(y = "", x = "P(r = 0)", title = "Implied prior CDF on P(r = 0)", color = "E[r | r > 0]")
```

This shows that the prior on $\phi_1$ expresses a lot of uncertainty about $P(r = 0)$
with high sequencing depth.

In the rest of the document, when we specify values of $\phi_1$ for numerical examples,
we'll look at three values: 0.1, 0.25, and 0.5.
These are close to the 15th, 50th, and 85th percentiles of the prior
and span $P(0)$ for deep sequencing of $\approx 7-54\%$:
```{r}
df_phi %>%
  filter(phi %in% c(0.1, 0.25, 0.5), r == 100) %>%
  select(phi, p0, prior) %>%
  rename(phi_1 = phi, `P(0)` = p0, `Prior CDF` = prior) %>%
  mutate(across(everything(), ~round(.x, digits = 2))) %>%
  kable

df_phi %>% filter(r == 100)
```

# The MLE

## Prevalence

To find the MLE prevalence $\hat{p}$, we maximize $\log L$ by setting its $p$ derivative to zero:

$$\frac{\partial \log L}{\partial p} = -n_0s(1 - P(0|\lambda, \phi_1)) + \frac{n_{>0}}{p} = 0$$
Solving for $\hat{p}$ gives:
$$\hat{p}(1 - P(0|\lambda, \phi_1)) = \frac{n_{>0}}{n_0 s}.$$

This has a nice interpretation: the MLE prevalence times the probability that
a positive swab contributes reads to a pool equals the number of pools with viral reads
divided by the number of swabs in pools with no viral reads.
(Why isn't the denominator of the RHS the total number of swabs?
Maybe it has to do with our approximation; maybe it has to do with avoiding double-counting.
Either way, the relative difference should be on the order of $sp(1 - P(0))$ and so consistent with our approximation.)

If the probability of getting zero reads from a positive swab is high---either because
the average number of viral reads from a positive swab ($\lambda$) is small
or because the overdispersion is large ($\phi_1$ is small)---then $\hat{p}$ must be larger to compensate.
For example, if $\lambda = \phi_1 = 1$, $P(0) = \frac{1}{2}$ and $\hat{p}$ must be twice as large
as if $P(0) = 0$.

## Expected viral reads

Following the same procedure for $\lambda$, we get
$$
\frac{\partial \log L}{\partial \lambda} =
n_0 s p \frac{\partial}{\partial \lambda} P(0) + \sum_{i:r_i > 0} \frac{\partial}{\partial \lambda} \log P(r_i) = 0,
$$

which we can solve for $\hat{\lambda}$ using the NB model (@eq-nb) for $P(r)$:
$$
\frac{\hat{\lambda}}{1 - P(0|\hat{\lambda}, \phi_1)} = \frac{\sum_i r_i}{n_{>0}}.
$$

As with $\hat{p}$, this has a nice interpretation.
The left-hand side is the expected number of viral reads in a pool conditional on there being any at all.
The right-hand side is the total number of viral reads divided by the number of pools with viral reads.

## Numerical examples

For given data and $\phi_1$, we can solve for $\hat{\lambda}$ and then for 
$P(0 | \hat{\lambda}, \phi_1)$ and $\hat{p}$.
