---
title: "Appendix to"
subtitle: "[Estimating the sensitivity of wastewater metagenomic sequencing using nasal swabs](https://naobservatory.org/blog/swab-based-p2ra/)"
format:
  html:
    code-fold: true
    toc: true
execute:
  cache: false
knitr:
  opts_chunk: 
    message: false
editor: source 
author: 
  - Dan Rice
  - Simon Grimm
  - Mike McLaren
date: 2025-06-06
date-modified: today
---

For bioinformatics and statistical analysis code, see the project's [GitHub repo](https://github.com/naobservatory/swab-based-p2ra).

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(rstan)
library(knitr)
library(here)
library(kableExtra)

theme_set(theme_minimal())

source("../scripts/prepare_data.R")
source("../scripts/define_model.R")

# Colors borrowed from https://github.com/JLSteenwyk/ggpubfigs
wong_eight <- c(
  "#E69F00",
  "#56B4E9",
  "#009E73",
  "#F0E442",
  "#0072B2",
  "#D55E00",
  "#CC79A7",
  "#000000"
)
options(
  ggplot2.discrete.colour = function()
    scale_colour_manual(values = wong_eight)
)
options(
  ggplot2.discrete.fill = function()
    scale_fill_manual(values = wong_eight)
)

# For axis labels
scientific <- function(x) {
  parse(text = sprintf("10^{%f}", log10(x)))
}

# Helper function for plotting priors
# If the x-axis is log10-transformed,transform the density function of x to
# the density of log10(x) by a Jacobian adjustment:
# dividing by $\frac{d log_{10}(x){dx} = \frac{1}{x \log 10}$
# so that the areas under the curve give probabilities correctly
prior_density <- function(x, density_function, log10x = FALSE){
  if (log10x){
    data <- tibble(
      x = x,
      density = density_function(x) * x * log(10), # Jacobian adjustment
    )
  } else {
    data <- tibble(
      x = x,
      density = density_function(x),
    )
  }
  geom_line(
    aes(x, density),
    data = data,
    color="grey",
    linetype="dashed",
    inherit.aes = FALSE
    )
}

```



# Model

We define a Bayesian model to jointly estimate viral prevalence and $RA_p(1\%)$
from sequencing of pooled swab samples and wastewater samples.
Joint estimation allows us to propagate our uncertainty in the prevalence
to uncertainty in $RA_p(1\%)$.

We designed the model to account for the following factors:

1. We do not know the true number of swabs from infected people in each pool.
2. We do not know the sensitivity of our swab sequencing to detect each virus,
but we expect it to increase with sequencing depth.
3. We expect the viral read counts to be overdispersed relative to pure counting (Poisson) noise.
The amount of overdispersion is likely differ between swabs and wastewater.

For each virus, our model takes as input data the total read depth and number of
viral reads observed in each swab pool and wastewater sample, and the number of
swabs in each swab pool.
In overview:

- Each swab in each pool comes from an infected person
  with probability equal to the prevalence.
- Each swab from an infected person contributes a random number of viral reads
  to its pool drawn from a negative binomial distribution.
  The average number of viral reads contributed is proportional to the
  sequencing depth and inversely proportional to the number of swabs in the pool.
- Each wastewater sample has a random number of viral reads drawn from a negative
  binomial distribution. The average number of viral reads is proportional to the
  prevalence and the sequencing depth.

In the following sections, we give the technical definition of the model and
the prior distributions of all parameters, then describe our model fitting and
checking procedures.
  
## Swab pool read counts

Let $p$ be the prevalence of a virus.
For each swab pool $i$, let $s_i$ be the number of swabs and $s_{+,i}$ be the
number of swabs from infected people.
Assuming that each person has an independent probability of being infected,
$$s_{+, i} \sim \text{Binomial}(s_i, p)$$

Let $v_{swab,i}$ be the number of viral reads in swab pool $i$.
We assume that the viral reads from different swabs are independent and combine linearly so that
$$\mathbb{E}[v_{swab,i} | s_{+,i}] = r_i \, \frac{s_{+,i}}{s_i} \, \mu_{swab} \equiv \lambda_i,$$
where $r_i$ is the total read depth in pool $i$ and $\mu_{swab}$ is
a relative abundance factor equal to the the expected relative abundance of the virus in an infected swab.
Note that the expected number of viral reads is proportional to the fraction of swabs from infected people.

To account for counting noise as well as overdispersion from, e.g., variation
in swab quality, we use a negative binomial distribution for viral read counts.
Using [Stan's (mean, overdispersion) parameterization for the negative binomial](https://mc-stan.org/docs/functions-reference/unbounded_discrete_distributions.html#nbalt),
$$v_{swab, i} | s_{+,i} \sim \text{NegativeBinomial}(\lambda_i , s_{+,i} \, \phi_{swab}).$$
Here, $\phi_{swab}$ is the inverse-overdispersion parameter for when one swab is from
an infected person.
The smaller $\phi_{swab}$ is, the more overdispersed the read count distribution
is relative to Poisson.
(The factor of $s_{+,i}$ multiplying $\phi_{swab}$ is a consequence of adding
reads from different swabs.)
In the limit where read depth is high so that $\lambda_i \gg s_{+,i}\phi_{swab}$, 
the coefficient of variation of $v_{swab,i}$ is given by
$$CV \approx \frac{1}{\sqrt{s_{+,i} \, \phi_{swab}}}.$$
Thus, our model allows for substantial noise from low-quality swabs and other sources,
even when the sequencing depth is high.

## Wastewater read counts

We also use a negative binomial distribution for our wastewater reads.
For wastewater sample $j$, the number of viral reads is
$$v_{ww, i} \sim \text{NegativeBinomial}(r_j \, p \, \mu_{ww}, \phi_{ww})$$
where $r_j$ is the total read depth,
$p$ is the same prevalence as for swab samples,
$\mu_{ww}$ is a relative abundance factor, and
$\phi_{ww}$ is the inverse-overdispersion parameter for wastewater sequencing.
The relative abundance factor $\mu_{ww}$ is defined so that
the expected relative abundance in a wastewater sample, when the prevalence is $p$,
is given by $p \cdot \mu_{ww}$. 
Thus, in this parameterization, $RA_p(1\%) = 0.01 \cdot \mu_{ww}$.

# Prior Distributions

Our model has five parameters: one for the prevalence, two for the expected
relative abundance factors in the two sample sources, and two for the read
count overdispersion.
We chose prior distributions that reflect several orders of magnitude of
uncertainty in these parameters.
We confirmed, using prior predictive checks, that the observed number of samples with
zero reads, the largest viral read count, and the total viral read count
for each virus are within the probable ranges under the priors.

## Prevalence

The prevalence of respiratory viruses is both variable and unknown,
so we put a prior on $p$ that spans several orders of magnitude.

The left tail of our prior controls the probability that a virus is too rare to observe in our swab sampling.
We designed our model to apply to all viruses found in wastewater
(though we subsequenctly restricted our analysis to those with high swab read counts).
Because we observed several viruses with zero viral reads in swab data---including a number of respiratory viruses that we did not see in our wastewater data---we chose a prior that puts significant probability on prevalence below our swab sampling detection limit.
We collected 547 total swabs, so our best-case detection limit is roughly 0.2%.
(The exact shape of the left tail of the prior should not 
affect our prevalence estimates for viruses observed in at least one swab
sample because the likelihood function will sharply constrain the posterior
probability of prevalence below 0.2%.)

The right tail of the prior sets the upper bound
of our prevalence estimates when our sensitivity is low.
We assume that it's unlikely, but not impossible, that the prevalence of a single
respiratory virus is much above 10%.

With these considerations as guides, 
we put a logit-Normal prior on $p$:
$$\text{logit}(p) \sim \text{Normal}(-6, 2).$$
With this distribution, which constrains the values to be between zero and one, the prior median is 0.25%, the 85th percentile is 1.8%
and the 95th percentile is 12%.

```{r}
#| label: prior-p
#| fig-cap: "**Prior distribution of prevalence.** Note the logistic spacing of the x-axis."

prior_logit_p <- function(x) dnorm(x, mean_logit_p, sd_logit_p)

tibble(
  logit_p = seq(-14, 2, length.out = 1000),
  p = plogis(logit_p),
  prior = prior_logit_p(logit_p),
) %>%
  ggplot(aes(p, prior)) +
  geom_line() +
  scale_x_continuous(
    transform = "logit",
    breaks = c(10^seq(-6, -1, 1), 0.5),
  ) +
  labs(y = "", x = "")
```

This prior is broad enough to permit
the true prevalence to be much higher than the pool positivity rate would suggest,
allowing for significant false negative rate.
For example, the coronavirus or rhinovirus with the fewest swab reads is Rhinovirus A with nine reads across three pools.
Assuming perfect sensitivity, the maximum likelihood prevalence for Rhinovirus A would be 0.7%.
Our prior assigns a 9% probability that the true prevalence is more than five times higher than this, i.e., a false negative rate over 83%.

## Relative abundance factors

For each sample type, the expected relative abundance is the product of the prevalence
and a relative abundance factor $\mu$.
For wastewater sequencing, the expected relative abundance at prevalence $p$ is
$p\, \mu_{ww}$, so that $RA_p(1\%) = 0.01 \mu_{ww}$.
For pooled swabs, the expected relative abundance is $p \, \mu_{swab}$,
where $\mu_{swab}$ is the expected relative abundance in a swab from an infected person.
Since these parameters represent proportions we use logit-normal priors to constrain their values between zero and one, as with the prevalence.

Because we do not have much prior information about the relative abundance factors,
we chose priors that span many orders of magnitude.
Our previous estimates for Influenza A and SARS-CoV-2 in our wastewater data suggest a range
of $\mu_{ww}$ from 1e-8 to 1e-4.
We would like to include more uncertainty than this range, so we use:

$$\text{logit}(\mu_{ww}) \sim \text{Normal}(-14, 6)$$
With these parameters, the median $\mu_{ww} is 8e-7 compared to our previous Flu A estimate of 1e-6.
The prior allows for a lot of uncertainty: 96% of the probability is between 5e-12 and 0.1,
spanning more than 10 orders of magnitude.

Because we are studying respiratory viruses,
we expect swabs to have higher relative abundance than wastewater at the same prevalence.
Accordingly, we'll shift the prior for $\mu_{swab}$ to the right relative to $\mu_{ww}$.
However, we don't want to make this assumption too strongly,
so we'll shift the mean by less than a standard deviation (keeping the standard deviation constant) to:
$$\text{logit}(\mu_{swab}) \sim \text{Normal}(-10, 6).$$
With these priors, the median and geometric mean of $\frac{\mu_{swab}}{\mu_{ww}} \approx 60$
and the probability that a given virus's $\mu_{swab} > \mu_{ww}$ is $\approx 70\%$.

```{r}
#| label: prior-mu
#| fig-cap: "**Prior distibutions of relative abundance factors $\\mu$.** Note the logistic spacing of the x-axis."

prior_logit_mu_swab <- function(x) dnorm(x, mean_logit_mu_swab, sd_logit_mu_swab)
prior_logit_mu_ww <- function(x) dnorm(x, mean_logit_mu_ww, sd_logit_mu_ww)

crossing(
  source = factor(c("swab", "wastewater")),
  logit_mu = seq(-31, 5, length.out = 1000),
  ) %>%
  mutate(
    mu = plogis(logit_mu),
    prior = if_else(
      source == "swab",
      prior_logit_mu_swab(logit_mu),
      prior_logit_mu_ww(logit_mu)
    ),
  ) %>%
  ggplot(aes(mu, prior, linetype = source)) +
  geom_line() +
  scale_x_continuous(
    transform = "logit",
    breaks = c(10^seq(-12, -2, 2), 0.5), #, 0.99),
  ) +
  labs(x="", y="")
```


## Overdispersion

Finally, we have the inverse-overdispersion parameters for the read counts:
$\phi_{swab}$ and $\phi_{ww}$.
We have little prior information about these parameters except that they must be greater than zero,
and we expect overdispersion relative to Poisson in both sample sources.
Unfortunately, with only twelve samples of each type, we do not expect our data
to give us much information about $\phi$ either.
(This is born out: our posterior distributions are close to the priors, especially for $\phi_{swab}$, see below.)

For wastewater, we use a standard weakly informative prior for non-negative dimensionless quantities:
$$\phi_{ww} \sim \text{Gamma}(1, 1) = \text{Exponential}(1).$$
When viral read counts are large, their coefficient of variation is approximately $\frac{1}{\sqrt{\phi_{ww}}}$.
Our prior allows the CV to be either larger than one ($\phi_{ww} < 1$) or smaller than one ($\phi_{ww} > 1$).

In practice, $\phi_{ww}$ does not appear to have much effect on our inference of $RA_p(1\%)$.
Plotting the the joint posterior distributions of $\mu_{ww}$ and $\phi_{ww}$
shows them to be approximately independent (not shown).

In contrast, $\phi_{swab}$ is important for our inference of prevalence (and thus $RA_p(1\%))$.
In our model, $\phi_{swab}$ determines the false-negative rate when the read depth is high.
When one swab comes from an infected person,
the probability of getting zero viral reads in our negative binomial model is 
$$P(v = 0 | s_+ = 1) = {\left(\frac{\phi_{swab}}{\lambda + \phi_{swab}}\right)}^{\phi_{swab}},$$ {#eq-p0}
where $\lambda$ is the expected number of viral reads.
If $\phi_{swab} < 1$, this probability can be substantial even for large $\lambda$.
The higher the false positive rate, the larger $p$ has to be to account for the same data.

We expect that one major source of false negatives is variation in how well subjects swab themselves.
Poor swabs can contribute zero viral reads, even with deep sequencing.
Accordingly, we selected a prior that expresses a high degree of uncertainty
about the probability of getting zero reads from an infected swab at high sequencing depth:
$$\phi_{swab} \sim \text{Gamma}(2, 7),$$
which looks like:

```{r}
#| label: prior-phi
#| fig-cap: "**Prior distributions of inverse-overdispersion $\\phi$**"

prior_phi_swab <- function(x) dgamma(x, shape_phi_swab, rate_phi_swab)
prior_phi_ww <- function(x) dgamma(x, shape_phi_ww, rate_phi_ww)

crossing(
  source = factor(c("swab", "wastewater")),
  phi = seq(0.01, 10, 0.001)
) %>%
  mutate(
    prior = if_else(source == "swab", prior_phi_swab(phi), prior_phi_ww(phi)),
    # Jacobian adjustment so that area under the curve gives probabilities
    # when the x-scale is log10 transformed (see setup block for explaination)
    prior_log10 = prior * phi * log(10),
  ) %>%
  ggplot(aes(phi, prior_log10, linetype = source)) +
  geom_line() +
  geom_vline(xintercept = c(0.15, 0.28, 0.4), linetype = "dotted", color = "grey") +
  scale_x_log10() +
  labs(y = "", x = "")
```

We can use @eq-p0 to calculate the implied probability of false negatives for a given sequencing depth and relative abundance factor.
As a proxy for "deep sequencing", assume that we sequence deeply enough that we expect to find 100 viral reads conditional on finding at least one.
Then, the dotted grey lines show the values of $\phi_{swab}$ that imply the probability of
getting zero viral reads from a pool containing one swab from an infected person is (from left to right) 0.4, 0.2, and 0.1.
The prior thus allows for a broad range of false negative rates.

# Model fitting

We implemented and fit the model using the [RStan](https://mc-stan.org/rstan/) interface to the [Stan](https://mc-stan.org/) programming language.
We fit the model independently to each virus observed in the wastewater data.
For each virus, we ran four chains for 4,000 iterations each with a warm-up of 1,000 iterations.

After fitting, we checked for convergence problems, which indicate
that the program had difficulty sampling from the posterior distribution.
For all viruses, including those with few or no swab reads, we confirmed that the chains mixed well ([rhat](https://mc-stan.org/learn-stan/diagnostics-warnings.html#r-hat) < 1.005) and that there were no [divergent transitions](https://mc-stan.org/learn-stan/diagnostics-warnings.html#divergent-transitions-after-warmup).

For instructions to reproduce the model fits and diagnostics, see
the project's [git repository](https://github.com/naobservatory/swab-based-p2ra).

# Model checking

## Posterior predictive checks

We performed posterior predictive checks to test that the fitted model captures
the general features of the data.
For each posterior sample of the parameters, we simulated a dataset under the model.
For each virus and sample source, we calculated the posterior predictive distribution of

1. The number of samples containing viral reads
2. The total viral reads
3. The maximum viral reads in a sample
4. The standard deviation of viral reads.

We visually compared these distributions to the observed quantities.
We did not find any large or systematic deviations from the posterior predictive distributions,
indicating that the model predicts these features of the data well.

## Posterior distributions

We examined our posterior distributions to understand the influence
of our priors and the structure of our model on our inferences
of prevalence and $\mu_{ww}$ (and thus $RA_p(1\%)$).

### The influence of the prior distributions of prevalence and $\mu_{ww}$ is weak

We intended our priors on $p$ and $\mu_{ww}$ to be weakly informative
so that our data would constrain the posterior distributions relative
to the priors.

For coronaviruses and rhinoviruses,
the posterior distributions of these parameters are narrow compared to their priors:

```{r}
#| label: post-p
#| fig-cap: "**Posterior (solid) and prior (dashed) distributions of prevalence**"
#| warning: false

posteriors <- read_tsv(here(tables_dir, "posteriors.tsv")) %>%
  filter(group %in% c("Coronaviruses (SARS-CoV-2)", "Coronaviruses (seasonal)", "Rhinoviruses"))

# To get a logit-spaced x-axis for both the prior and the posterior,
# Plot logit p on the x-axis, but change the x-tick labels
# to p on the natural scale
breaks <- 10^seq(-4, -1, 1)
posteriors %>%
  ggplot(aes(logit_p, color=group)) +
  geom_density() +
  prior_density(seq(-12, 4, length.out=100), prior_logit_p) +
  scale_x_continuous(
    limits = qlogis(c(1e-5, 2e-1)),
    breaks = qlogis(breaks),
    labels = scientific(breaks),
    ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(
    transform = "log10",
    limits = c(1e-2, 1),
    breaks = 10^seq(0,-2),
    labels = scientific
  ) +
  labs(x = "", y = "") +
  theme(legend.position = "")
```

```{r}
#| fig-cap: "**Posterior (solid) and prior (dashed) distributions of $\\mu_{ww}$**"
#| warning: false
#| label: post-mu

# Plot logit mu on the x-axis, but change the x-tick labels to mu on the
# untransformed scale
breaks <- 10^seq(-7, -3, 2)
posteriors %>%
  ggplot(aes(logit_mu_ww, color = group)) +
  geom_density() +
  prior_density(seq(-20, -5, length.out = 100), prior_logit_mu_ww) +
  scale_x_continuous(
    limits = c(-20, -5),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "") +
  theme(legend.position = "")
```

This suggests that the data, interpreted in light of our model and the priors on the other parameters, effectively constrains the posteriors of these parameters.

For viruses with zero or one read in swab samples (not shown), the posterior distributions were dominated by the priors
because the model was not able to constrain the false negative rate of swab sequencing.

### Uncertainty in prevalence propagates to uncertainty in $\mu_{ww}$

Because the expected relative abundance of viral reads in wastewater is the product of $p$ and $\mu_{ww}$, we expect uncertainty in prevalence to contribute to uncertainty in $\mu_{ww}$ (and thus $RA_p(1\%)$).

Indeed, posterior uncertainty in $p$ explains much of the posterior uncertainty in $\mu_{ww}$.
Plotting the joint posterior distribution of these parameters shows a tight correlation:

```{r}
#| fig-cap: "**Posterior samples of $\\mu_{ww}$ vs prevalence**"
posteriors %>%
  ggplot(aes(p, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species, ncol = 4) +
  labs(x = "", y = "")
```

We can calculate the posterior fraction of variance in $\log \mu_{ww}$ explained by $\log p$ ($R^2$):

```{r}
#| fig-cap: Fraction of posterior variance in $\log \mu_{ww}$ explained by $\log p$
posteriors %>%
  summarise(
    cor_log = cor(log(p), log(mu_ww), method = "pearson"),
    R_squared = cor_log ^2,
    .by = species,
  ) %>%
  select(-cor_log) %>%
  kable(digits = c(NA, 2), col.names = c("species", "R<sup>2</sup>"))
```

For all viruses, at least 80% of the posterior variance in $\log \mu_ww$ is explained by variance in $\log p$.

### Uncertainty in the true positive rate contributes to uncertainty in prevalence

Much of our posterior uncertainty in the prevalence comes from our uncertainty
about the true positive rate of swab sequencing:
the probability that a swab pool will have at least one viral read when
it contains at least one swab from an infected person.
This probability, which is one minus the false negative rate, is controlled by the parameters $\mu_{swab}$ and $\phi_{swab}$.

For the coronaviruses and rhinoviruses, we observe enough reads to estimate $\mu_{swab}$:

```{r}
#| warning: false
# Plot logit mu on the x-axis, but change the x-tick labels to mu on the
# untransformed scale
breaks <- 10^seq(-5, -1, 2)
posteriors %>%
  ggplot(aes(logit_mu_swab, color = group)) +
  geom_density() +
  prior_density(seq(-16, 2, length.out = 100), prior_logit_mu_swab) +
  scale_x_continuous(
    limits = c(-14, 2),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "mu_swab") +
  theme(legend.position = "")
```

On the other hand, the data provide very little information about $\phi_{swab}$ relative to the priors:

```{r}
#| warning: false
posteriors %>%
  ggplot(aes(phi_swab, color=group)) +
  geom_density() +
  prior_density(10^seq(-3, 3, length.out=100), prior_phi_swab, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1)) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 2)) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "")
```

Because the data does not constrain $\phi_{swab}$ very much,
our prior $\phi_{swab}$ has a meaningful impact on our estimates.

To see the combined effect of $\mu_{swab}$ and $\phi_{swab}$ on our prevalence estimates,
we can transform the posterior samples of $\mu_{swab}$ and $\phi_{swab}$ into the
posterior samples of the true positive rate:

$$\text{true positive rate} \approx 1 - P(v = 0 | s_+ = 1), $$
where we make the approximation that most pools have either zero or one swab from an infected person.
$P(v = 0 | s_+ = 1)$ depends on $\phi_{swab}$, $\mu_{swab}$, the total read depth and the number of swabs in the pool according to @eq-p0.
Because our pools vary in number of swabs and read depth, we average the true positive rate over pools.

```{r}
#| warning: false
#| fig-cap: "**Posterior prevalence vs posterior true positive rate**"

# For each posterior sample and each swab pool, calculate the probability of
# getting zero reads if a single swab comes from an infected person.
# Then estimate the true positive rate in each posterior sample
# as the average across pools of 1 - p(0).
true_positive_rate <- posteriors %>%
  mutate(posterior_sample = row_number()) %>%
  crossing(swab_metadata) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    tpr = 1 - mean(p0),
    .by = c(species, group, posterior_sample),
  )

true_positive_rate %>%
  ggplot(aes(tpr, p)) +
  geom_point(alpha = 0.01, size = 0.1) +
  scale_x_log10(limits = c(8e-3, 1.2)) +
  scale_y_log10() +
  facet_wrap(~species, ncol = 4) +
  theme(legend.position="none") +
  labs(y = "", x = "")
```

```{r}
#| fig-cap: $R^2$ of $\log p$ vs $\log \text{true positive rate}$
true_positive_rate %>%
  summarise(
    cor_log = cor(log(tpr), log(p), method = "pearson"),
    R_squared = cor_log ^2,
    .by = species,
  ) %>%
  select(-cor_log) %>%
  kable(digits = c(NA, 2), col.names = c("species", "R<sup>2</sup>"))
```

When the true positive rate is lower (i.e., the false negative rate is higher),
the estimated prevalence is higher to account for the observed positive pools.
The median R-squared between log posterior prevalence and log posterior true positive rate is 0.3, implying that about a third of the uncertainty in prevalence comes from uncertainty in the true positive rate.
