---
title: "[WIP] Appendix"
format:
  html:
    code-fold: true
    toc: true
  pdf:
    toc: true
knitr:
  opts_chunk: 
    message: false
editor: source 
author: dan
date: 2025-05-30
date-modified: today
---


```{r setup}
#| echo: false

library(tidyverse)
library(rstan)
library(knitr)
library(kableExtra)
theme_set(theme_minimal())

# Colors borrowed from https://github.com/JLSteenwyk/ggpubfigs
wong_eight <- c(
  "#E69F00",
  "#56B4E9",
  "#009E73",
  "#F0E442",
  "#0072B2",
  "#D55E00",
  "#CC79A7",
  "#000000"
)
options(
  ggplot2.discrete.colour = function()
    scale_colour_manual(values = wong_eight)
)
options(
  ggplot2.discrete.fill = function()
    scale_fill_manual(values = wong_eight)
)

options(mc.cores = parallel::detectCores())

# For axis labels
scientific <- function(x) {
  parse(text = sprintf("10^{%f}", log10(x)))
}

# Helper function for plotting priors
prior_density <- function(x, density_function, log10x = FALSE){
  if (log10x){
    data <- tibble(
      x = x,
      density = density_function(x) * x * log(10), # Jacobian adjustment
    )
  } else {
    data <- tibble(
      x = x,
      density = density_function(x),
    )
  }
  geom_line(
    aes(x, density),
    data = data,
    color="grey",
    linetype="dashed",
    inherit.aes = FALSE
    )
}
```

```{r}
swab_metadata <- read_tsv("../../tables/swab-sample-metadata.tsv") %>%
  mutate(date = ymd(paste0("20", date)))

# Use the per-day data. Caveat about the different treatments
swab_reads <- read_tsv("../../tables/swabs-ra-summary.tsv") %>%
  mutate(date = ymd(paste0("20", date))) %>%
  select(-all_reads)

wastewater_metadata <- read_tsv("../../tables/wastewater-sample-metadata.tsv") %>%
  mutate(date = ymd(paste0("20", date)))

wastewater_reads <- read_tsv("../../tables/ww-ra-summary.tsv") %>%
  mutate(date = ymd(paste0("20", date))) %>%
  select(-all_reads)

all_species <- bind_rows(swab_reads, wastewater_reads) %>%
  select(species, group) %>%
  distinct %>%
  arrange(group, species)

species_order <- all_species$species


complete_data <- function(all_species, metadata, read_data){
  complete <- crossing(all_species, metadata) %>%
    left_join(
      summarise(
        read_data,
        .by = c(date, location, species, group),
        viral_reads = sum(dedup_hv)
      ),
      by = join_by(date, location, species, group),
    ) %>%
    arrange(group, species) %>%
    mutate(
      viral_reads = replace_na(viral_reads, 0),
      species = fct_inorder(species),
      group = fct_inorder(group),
      )
}

all_species <- bind_rows(swab_reads, wastewater_reads) %>%
  select(species, group) %>%
  distinct %>%
  arrange(group, species)

swab_reads_complete <- complete_data(
  all_species,
  swab_metadata,
  swab_reads
  )
wastewater_reads_complete <- complete_data(
  all_species,
  wastewater_metadata,
  wastewater_reads
  )
```


# Model

We define a Bayesian model to jointly estimate viral prevalence and $RA_p(1\%)$
from sequencing of pooled swab samples and wastewater samples.
Joint estimation allows us to propagate our uncertainty in the prevalence
to uncertainty in $RA_p(1\%)$.

We designed the model to account for the following factors:

1. We do not know the true number of swabs from infected people in each pool.
2. We do not know the sensitivity of our swab sequencing to detect each virus,
but we expect it to increase with sequencing depth.
3. We expect the viral read counts to be overdispersed relative to pure counting (Poisson) noise.
The amount of overdispersion is likely to be different between swabs and wastewater.

## Swab pool read counts

Let $p$ be the prevalence of a virus.
For each swab pool $i$, let $s_i$ be the number of swabs and $s_{+,i}$ be the
number of swabs from infected people.
Assuming that each person has an independent probability of being infected,
$$s_{+, i} \sim \text{Binomial}(s_i, p)$$

Let $v_{swab,i}$ be the number of viral reads in swab pool $i$.
We assume that the viral reads from different swabs are independent and combine linearly so that
$$\mathbb{E}[v_{swab,i} | s_{+,i}] = r_i \, \frac{s_{+,i}}{s_i} \, \mu_{swab} \equiv \lambda_i,$$
where $r_i$ is the total read depth in pool $i$ and $\mu_{swab}$ is the expected
relative of the virus in an infected swab.
Note that the expected number of viral reads is proportional to the fraction of swabs
from infected people.

To account for counting noise as well as overdispersion from, e.g., variation
in swab quality, we use a negative binomial distribution for viral read counts.
Using [Stan's (mean, overdispersion) parameterization for the negative binomial](https://mc-stan.org/docs/functions-reference/unbounded_discrete_distributions.html#nbalt),
$$v_{swab, i} | s_{+,i} \sim \text{NegativeBinomial}(\lambda_i , s_{+,i} \, \phi_{swab}).$$
Here, $\phi_{swab}$ is the inverse-overdispersion parameter for when one swab is from
an infected person.
The smaller $\phi_{swab}$ is, the more overdispersed the read count distribution
is relative to Poisson.
(The factor of $s_{+,i}$ multiplying $\phi_{swab}$ is a consequence of adding
reads from different swabs.)
In the limit where read depth is high so that $\lambda_i \gg s_{+,i}\phi_{swab}$, 
the coefficient of variation of $v_{swab,i}$ is given by
$$CV \approx \frac{1}{\sqrt{s_{+,i} \, \phi_{swab}}}.$$
Thus, our model allows for substantial noise from swab quality,
even when the sequencing depth is high.

## Wastewater read counts

We also use a negative binomial distribution for our wastewater reads.
For wastewater sample $j$, the number of viral reads is
$$v_{ww, i} \sim \text{NegativeBinomial}(r_j \, p \, \mu_{ww}, \phi_{ww})$$
where $\phi_{ww}$ is the inverse-overdispersion parameter for wastewater sequencing.
The expected relative abundance in a wastewater sample, when the prevalence is $p$,
is given by $p \cdot \mu_{ww}$. 
Thus, in this parameterization, $RA_p(1\%) = 0.01 \cdot \mu_{ww}$.

# Prior Distributions

Our model has five parameters: one for the prevalence, two for the expected
relative abundance in the two sample sources, and two for the read
count overdispersion.
We chose prior distributions that reflect several orders of magnitude of
uncertainty in these parameters.
We confirmed, using prior predictive checks, that the number of samples with
zero reads, the maximum read counts, and the total read counts for each virus
are within the probable ranges under the priors.

## Prevalence

The prevalence of respiratory viruses is both variable and unknown,
so we put a prior on $p$ that spans several orders of magnitude.

We collected 547 total swabs, so our detection limit is roughly 0.2%.
Because we had several viruses with zero viral reads in swab data---including a number of respiratory viruses that we did not see in our wastewater data---our prior should put significant probability on prevalence below this limit.
However, the exact shape of the left tail should not matter for viruses observed at least once.

The right tail of the prior is more important because it sets the upper bound
of our prevalence estimates when our sensitivity is low.
We assume that it's unlikely, but not impossible, that the prevalence of a single
respiratory virus is much above 10%.

With these considerations as guides, 
we put a logit-Normal prior on $p$, which constrains the values to be between zero and one:
$$\text{logit}(p) \sim \text{Normal}(-6, 2).$$
With this distribution, the prior median is 0.25%, the 85th percentile is 1.8%
and the 95th percentile is 12%.

```{r}
#| label: prior-p

mean_logit_p <- -6
sd_logit_p <- 2
prior_logit_p <- function(x) dnorm(x, mean_logit_p, sd_logit_p)

tibble(
  logit_p = seq(-14, 2, length.out = 1000),
  p = plogis(logit_p),
  prior = prior_logit_p(logit_p),
) %>%
  ggplot(aes(p, prior)) +
  geom_line() +
  scale_x_continuous(
    transform = "logit",
    breaks = c(10^seq(-6, -1, 1), 0.5),
    name = "prevalence (logistic scale)",
  ) +
  labs(title = "Prior on prevalence", y = "")
```

This prior allows for the possibility that the sensitivity of our swab sequencing is low and the true prevalence is higher than the pool positivity rate would suggest.
The Coronavirus or Rhinovirus with the fewest viral reads is Rhinovirus A with 9 reads across three pools.
Assuming perfect sensitivity, the maximum likelihood prevalence for Rhinovirus A would be 0.069%.
Our prior assigns a 9% probability that the true prevalence is more than five times higher than this.

## Expected relative abundance

For each sample type, the expected relative abundance is the product of the prevalence
and a parameter $\mu$.
For wastewater sequencing, the expected relative abundance at prevalence $p$ is
$p\, \mu_{ww}$, so that $RA_p(1\%) = 0.01 \mu_{ww}$.
For pooled swabs, the expected relative abundance is $p \, \mu_{swab}$,
where $\mu_{swab}$ is the expected relative abundance in a swab from an infected person.
Since these parameters represent proportions, we use logit-normal priors to constrain
their values between zero and one.

Because we do not have much prior information about these parameters,
we chose priors that span many orders of magnitude.
Our previous estimates for Flu and SARS-CoV-2 in wastewater suggest a range
of $\mu_{ww}$ from 1e-8 to 1e-4.
We would like to include more uncertainty than this range, so we use:

$$\text{logit}(\mu_{ww}) \sim \text{Normal}(-14, 6)$$
With these parameters:

 -  The median is 8.3e-7. This corresponds to an $RA_p(1\%)$ of 8.3e-9,
    compared to our previous Flu A estimate of 1.4e-8.
 -  70% of the probability is between 2e-9 and 3e-4.
 -  96% of the probability is between 5e-12 and 0.1,
    spanning more than 10 orders of magnitude.

Because we are studying respiratory viruses,
we might expect the swabs to have higher relative abundance than wastewater at the same prevalence.
Accordingly, we'll shift the prior for $\mu_{swab}$ to the right relative to $\mu_{ww}$.
However, we don't want to make this assumption too strongly,
so we'll shift it by less than a standard deviation to:
$$\text{logit}(\mu_{swab}) \sim \text{Normal}(-10, 6).$$
With these priors, the geometric mean of $\frac{\mu_{swab}}{\mu_{ww}} \approx 58$
and the probability that a given virus's $\mu_{swab} > \mu_{ww}$ is 68%.

```{r}
#| label: prior-mu

# swab
mean_logit_mu_swab <- -10  # Prior mean on logit scale
sd_logit_mu_swab <- 6      # Prior standard deviation on logit scale
prior_logit_mu_swab <- function(x) dnorm(x, mean_logit_mu_swab, sd_logit_mu_swab)

# wastewater
mean_logit_mu_ww <- -14  # Prior mean on logit scale
sd_logit_mu_ww <- 6      # Prior standard deviation on logit scale
prior_logit_mu_ww <- function(x) dnorm(x, mean_logit_mu_ww, sd_logit_mu_ww)

crossing(
  source = factor(c("swab", "wastewater")),
  logit_mu = seq(-31, 5, length.out = 1000),
  ) %>%
  mutate(
    mu = plogis(logit_mu),
    prior = if_else(
      source == "swab",
      prior_logit_mu_swab(logit_mu),
      prior_logit_mu_ww(logit_mu)
    ),
  ) %>%
  ggplot(aes(mu, prior, linetype = source)) +
  geom_line() +
  scale_x_continuous(
    transform = "logit",
    breaks = c(10^seq(-12, -2, 2), 0.5), #, 0.99),
    name = expression(mu~"(logistic scale)"),
  ) +
  labs(title = expression("Priors on "~mu))
```


## Overdispersion

Finally, we have the inverse-overdispersion parameters for the read counts:
$\phi_{swab}$ and $\phi_{ww}$.
We have little prior information about these parameters except that they must be greater than zero,
and we expect overdispersion relative to Poisson in both sample sources.
Unfortunately, with only twelve samples of each type, we do not expect our data
to give us much. information about $\phi$ either.
(This is born out: our posterior distributions are close to the priors, especially for $\phi_{swab}$.)

For wastewater, a weakly informative prior:
$$\phi_{ww} \sim \text{Gamma}(1, 1) = \text{Exponential}(1).$$
This prior allows for both strong overdispersion $\phi_{ww} < 1$ and weak overdispersion $\phi_{ww} > 1$.
In practice, $\phi_{ww}$ does not appear to have much effect on our inference of $RA_p(1\%)$.

For swabs, the most important impact of $\phi_{swab}$ on our estimates of prevalence
comes from the fact that it, along with $\mu_{swab}$, controls the probability of getting zero viral reads from a positive swab.
We expect that one major source of these false negatives is variation in how well subjects swab themselves.
Poor swabs can contribute zero viral reads, even with deep sequencing.

Accordingly, we selected a prior that expresses a high degree of uncertainty
about the probability of getting zero reads from an infected swab at high sequencing depth:
$$\phi_{swab} \sim \text{Gamma}(2, 7),$$
which looks like:

```{r}
#| label: prior-phi

shape_phi_swab <- 2
rate_phi_swab <- 7
prior_phi_swab <- function(x) dgamma(x, shape_phi_swab, rate_phi_swab)

shape_phi_ww <- 1
rate_phi_ww <- 1
prior_phi_ww <- function(x) dgamma(x, shape_phi_ww, rate_phi_ww)

crossing(
  source = factor(c("swab", "wastewater")),
  phi = seq(0.01, 10, 0.001)
) %>%
  mutate(
    prior = if_else(source == "swab", prior_phi_swab(phi), prior_phi_ww(phi)),
    prior_log10 = prior * phi * log(10), # Jacobian adjustment
  ) %>%
  ggplot(aes(phi, prior_log10, linetype = source)) +
  geom_line() +
  geom_vline(xintercept = c(0.15, 0.28, 0.4), linetype = "dotted", color = "grey") +
  scale_x_log10() +
  labs(title = expression("Priors on inverse-overdispersion"), y = "", x = expression(phi))
```

Dotted grey lines show the values of $\phi_{swab}$ that imply the probability of
getting zero viral reads from an infected person's swab is 0.4, 0.2, and 0.1 from left to right
(assuming that we expect to see 100 viral reads conditional on seeing any).

# Model fitting

We implemented and fit the model using the `rstan` interface to the [Stan](https://mc-stan.org/) programming language.
We fit the model independently to each virus observed in the wastewater data.
For each virus, we ran four chains for 4,000 iterations each with a warm-up of 1,000 iterations.

After fitting, we checked for convergence problems, which indicate
that the program had difficulty sampling from the posterior distribution.
We confirmed that the chains mixed well ([rhat](https://mc-stan.org/learn-stan/diagnostics-warnings.html#r-hat) < 1.005) and that there were no [divergent transitions](https://mc-stan.org/learn-stan/diagnostics-warnings.html#divergent-transitions-after-warmup).

# Model checking

## Posterior predictive checks

We performed posterior predictive checks to test that the fitted model captures
the general features of the data.
For each posterior sample of the parameters, we simulated a dataset under the model.
For each virus and sample source, we calculated the posterior predictive distribution of

1. The number of samples containing viral reads
2. The total viral reads
3. The maximum viral reads in a sample
4. The standard deviation of viral reads.

We compared these distributions to the observed quantities.
We did not find any large or systematic deviations from the posterior predictive distributions,
indicating that the model predicts these features of the data well.

## Posterior distributions

We compared our posterior distributions to our priors to check that the priors
are not inappropriately dominating the inference of prevalence and $\mu_{ww}$
(and thus $RA_p(1\%)$).

For the Coronaviruses and Rhinoviruses,
the posteriors are narrow compared to the priors, suggesting that most of the
information in the inference is coming from the data.

```{r}
#| fig-cap: "Posterior (solid) and prior (dashed) distributions of prevalence"
#| warning: false

posteriors <- read_tsv("../../tables/posteriors.tsv") %>%
  filter(group %in% c("Coronaviruses (SARS-CoV-2)", "Coronaviruses (seasonal)", "Rhinoviruses"))

posteriors %>%
  ggplot(aes(p, color=group)) +
  geom_density() +
  prior_density(plogis(seq(-12, 4, length.out=100)), compose(prior_logit_p, qlogis)) +
  scale_x_continuous(transform = "logit", limits = c(1e-5, 2e-1), breaks = 10^seq(-4, -1, 1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(
    transform = "log10",
    limits = c(1e-2, 1),
    breaks = 10^seq(0, -2),
    labels = scientific
  ) +
  labs(x = "", y = "") +
  theme(legend.position = "")
```

```{r}
#| fig-cap: "Posterior (solid) and prior (dashed) distributions of $\\mu_{ww}$"
#| warning: false

breaks <- 10^seq(-7, -3, 2)
posteriors %>%
  ggplot(aes(logit_mu_ww, color = group)) +
  geom_density() +
  prior_density(seq(-20, -5, length.out = 100), prior_logit_mu_ww) +
  scale_x_continuous(
    limits = c(-20, -5),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "") +
  theme(legend.position = "")
```

Much of our posterior uncertainty in the prevalence comes from our uncertainty
about the sensitivity of our swab sequencing to detect reads if an infected
person contributes a swab.
We can transform our posterior samples of $\mu_{swab}$ and $\phi_{swab}$ into the
posterior probability that we observe viral reads in a sample given that one infected person contributed a swab to the pool.
(Because the pools vary in number of swabs and total sequencing depth, we average this probability over the pools.)

```{r}
#| warning: false

posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  ggplot(aes(1 - mean_p0, p)) +
  geom_point(alpha = 0.01, size = 0.1) +
  scale_x_log10(limits = c(8e-3, 1.2)) +
  scale_y_log10() +
  facet_wrap(~species, ncol = 4) +
  theme(legend.position="none") +
  labs(y = "posterior prevalence", x = "Pr(viral reads > 0 | infected swab)")
```

We find that this measure of sensitivity explains much of our uncertainty in prevalence.
For viruses where we observed many viral reads (HCoV-229E and SARS-CoV-2),
the estimated sensitivity is higher and it explains less of the uncertainty in prevalence.

```{r}
knitr::knit_exit()
```

# END CURRENT VERSION

```{r}
species_swab_summary <- swab_reads_complete %>%
  summarise(
    pos_pools = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    .by = c(species, group),
  ) %>%
  arrange(species)
kable(species_swab_summary)
```

```{r}
species_ww_summary <- wastewater_reads_complete %>%
  summarise(
    pos_samples = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    .by = c(species, group),
  ) %>%
  arrange(species)
kable(species_ww_summary)
```


# Posterior distributions

## Marginal posterior distributions

Let's visualize the posterior distributions for our key parameters.
In the following plots, solid lines show the posterior distribution of prevalence for each virus species, compared to the prior distribution (dashed line).
Colors represent viral groups (SARS-CoV-2, seasonal Coronaviruses, Influenza, Mononegavirales, and Rhinoviruses).


### Expected viral fractions

Posteriors on $\mu_{swab}$ (swab viral fraction with all positive swabs).

```{r}
breaks <- 10^seq(-5, -1, 2)
posteriors %>%
  ggplot(aes(logit_mu_swab, color = group)) +
  geom_density() +
  prior_density(seq(-16, 2, length.out = 100), prior_logit_mu_swab) +
  scale_x_continuous(
    limits = c(-14, 2),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "mu_swab") +
  theme(legend.position = "")
```

For species with multiple positive pools, (e.g. the Rhinoviruses), there is information to constrain the expected viral fraction in a positive swab.
With only one positive pool (e.g. H1N1, which has only one viral read), there is less, but still some information.

Note that all of the posteriors are shifted to the right relative to the prior.
This means that the prior is biasing these estimates downward.
This is a predictable effect of using the same prior for $\mu_{swab}$ and $\mu{ww}$
We may want to consider having separate priors.

HCoV-229E is an outlier because it has only a single positive pool with 6008 viral reads.
This puts most of the weight on $\mu_{swab} > 0.1$, which would be higher if the prior weren't pulling it down.
I'm not sure whether this is a good effect: regularization, or a bias we'd like to eliminate.
It depends on whether we think it's likely that $\mu_{swab}$ could actually be this high, vs this being a very lucky sample.

### Overdispersion

Posteriors for swab inverse-overdispersion $\phi_{swab}$.

```{r}
posteriors %>%
  ggplot(aes(phi_swab, color=group)) +
  geom_density() +
  prior_density(10^seq(-3, 3, length.out=100), prior_phi_swab, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1)) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "", axis.text.y = element_blank())
```

The posterior distributions are all very close to the prior,
suggesting that, as expected the data contains very little information about $\phi_{swab}$, at least relative to the prior.

There is a slight leftward shift for HCoV-OC43, which has the most positive pools (5).

Posteriors for wastewater inverse-overdispersion $\phi_{ww}$.

```{r}
posteriors %>%
  ggplot(aes(phi_ww, color=group)) +
  geom_density() +
  prior_density(10^seq(-2, 3, length.out=100), prior_phi_ww, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "", axis.text.y = element_blank())
```

The wastewater data has more non-zero counts and thus more data to infer dispersion.
I'm not sure I trust any of these though.

Generally, the species that have high predicted overdispersion relative to the prior (low $\phi$) are ones with fewer positive wastewater samples. 
That could be because there are more zero-read samples than expected given the counts in the non-zero-read samples.
Or it could be because the estimates are noisier with less data.

## Joint Posterior Distributions

We can also examine relationships between different parameters in the posterior.
In the following plots, each point is a sample from the joint posterior of two parameters.
The curves are an estimate of the conditional mean of the y-axis parameter given the x-axis parameter.

### Realtionships with prevalence

Prevalence vs $\mu_{swab}$ (note log x-scale not logit):

```{r}
posteriors %>%
  ggplot(aes(mu_swab, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(limits = c(1e-7, 1), labels = scientific) +
  scale_y_log10(limits = c(1e-5, 0.5), labels = scientific) +
  facet_wrap(~species)
```

It looks like:

- For small $\mu$, prevalence grows as $\mu$ decreases.
- For larger $\mu$, this relationship flattens.

This makes sense because low $\mu$ requires high prevalence to observe the virus since each positive swab is likely to go unobserved, while with high $\mu$ positive swabs are usually observed, so the prevalence is determined by the number of positive pools.

TODO: Add maximum likelihood calculation here.

```{r}
posteriors %>%
  ggplot(aes(phi_swab, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(limits = c(1e-3, 10), labels=scientific) +
  scale_y_log10(limits = c(1e-5, 0.5), labels = scientific) +
  facet_wrap(~species)
```

Smaller $\phi_{swab}$ implies a higher probability of failing to get viral reads a positive swab.
Thus, the inferred prevalence conditional on $phi_{swab}$ is inversely related to $phi_{swab}$.

```{r}
posteriors %>%
  summarise(
    corr_log_p_phi_swab = cor(log(p), log(phi_swab), method = "spearman"),
    corr_p_phi_swab = cor(p, phi_swab, method = "spearman"),
    corr_p_mu_ww = cor(log(p), log(mu_swab), method = "spearman"),
    .by = species,
  )
```

To encapsulate the relationship between $p$, $\mu_{swab}$, and $\phi_{swab}$,
we can approximate $\hat{p}$, the maximum likelihood $p$ conditional on known $\mu_{swab}$ and
$\phi_{swab}$.
(The approximation is in the the limit where most pools are expected to be negative.)


```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  summarise(
    corr_sp = cor(log(mean_p0), log(p), method = "spearman"),
    corr_pe = cor(log(mean_p0), log(p), method = "pearson"),
    .by = c(species),
  )
```

The approximate maximum likelihood and posterior $p$ qualitatively agree.

### Relationships with expected wastewater viral fraction

Prevalence: 

```{r}
posteriors %>%
  ggplot(aes(p, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```

```{r}
posteriors %>%
  summarise(
    cor_sp = cor(p, mu_ww, method = "spearman"),
    cor_log = cor(log(p), log(mu_ww), method = "pearson"),
    .by = species,
  )
```

As expected, there is a very tight relationship between prevalence and estimated wastewater viral fraction (per prevalence).
Often, strong correlations between parameters suggest that we should reparameterize the model so the parameters are more independent.
However, in this case, I think this is appropriate. We need a distinct $p$ parameter to make use of the discreteness of the swab pools.

The next three plots show the relationship between $\mu_{ww}$ and $\mu_{swab}$, $\phi_{swab}$, and $\phi_{ww}$:

```{r}
posteriors %>%
  ggplot(aes(mu_swab, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```


```{r}
posteriors %>%
  ggplot(aes(phi_swab, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```

```{r}
posteriors %>%
  ggplot(aes(phi_ww, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10() + # label = scientific) +
  scale_y_log10(label = scientific) +
  facet_wrap(~species)
```

The relationships with $\mu_{swab}$ and $\phi_{swab}$ are mediated by $p$ and are not surprising.

The relationship with $\phi_{ww}$ suggests that the posteriors are uncorrelated,
which is encouraging because we don't have any real information about this parameter.

# Posterior Predictive Checks

## Swab data

To validate our model, we can use posterior predictive checks - generating new data from the fitted model and comparing it to the observed data.
The following table gives the summary statistics for the data.

```{r}
#| label: swab-ppd
#| cache: true

posterior_predictive_swab <- posteriors %>%
  mutate(rep = 1:n(), .by = species) %>%
  cross_join(swab_metadata) %>%
  mutate(
    n_pos = map2_int(pool_size, p, ~ rbinom(1, .x, .y)),
    mu = n_pos * mu_swab * all_reads / pool_size,
    phi = n_pos * phi_swab,
    viral_reads = map2_int(mu, phi, ~ ifelse(.x > 0, rnbinom(
      1, size = .y, mu = .x
    ), 0)),
    false_negative = n_pos > 0 & viral_reads == 0,
  )

posterior_predictive_swab_summary <- posterior_predictive_swab %>%
  summarise(
    pos_pools = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    false_negative_pools = sum(false_negative),
    across(c(p, mu_swab, phi_swab), first),
    .by = c(species, group, rep),
  )



kable(species_swab_summary)
```


```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = pos_pools, fill = group, after_stat(density))) +
  geom_histogram(binwidth = 1, center = 0, color = "grey") +
  geom_vline(data = species_swab_summary, mapping = aes(xintercept=pos_pools)) +
  facet_wrap(vars(species)) +
  scale_x_continuous(breaks = seq(0, 12, 2), name = "positive pools")
```

This plot compares the observed number of positive pools (vertical line) with the distribution of positive pools generated from our model.
For every virus, the observed number of positive pools is common in the simulations.

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_total, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_continuous(
    transform = "log10",
    breaks = 10^seq(0, 4),
    name = "total viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed total viral reads (points) with the distribution from our model.
Vertical bars show quartiles

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_max, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_continuous(
    # transform = "pseudo_log",
    # breaks = c(0, 10^seq(0, 4)),
    transform = "log10",
    breaks = 10^seq(0, 4),
    name = "largest viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed maximum viral reads with the distribution from our model.

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_sd, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_log10(
    breaks = 10^seq(0, 4),
    name = "stdev viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed standard deviation of viral reads with the distribution from our model.
Violins show only the simulations with at least one viral read.
Vertical bars show quartiles (conditional on at least one read).

Together, these plots suggest that the model is capturing the swab data reasonably well.

## Wastewater data

Now we'll perform similar posterior predictive checks for the wastewater data:

```{r}
#| label: wastewater-ppd
#| cache: true

posterior_predictive_ww <- posteriors %>%
  mutate(rep = 1:n(), .by=species) %>%
  cross_join(wastewater_metadata) %>%
  mutate(
    mu = p * mu_ww * all_reads,
    viral_reads = map2_int(mu, phi_ww, ~ ifelse(.x > 0, rnbinom(1, size = .y, mu = .x), 0)), 
  )

posterior_predictive_ww_summary <- posterior_predictive_ww %>%
  summarise(
    pos_samples = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    across(c(p, mu_ww, phi_ww), first),
    viral_reads_expected = p * mu_ww * sum(all_reads),
    .by = c(species, group, rep),
  )


kable(species_ww_summary)
```

This code calculates summary statistics for the observed wastewater data, which we'll use to compare with predictions from our model.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = pos_samples, fill = group, after_stat(density))) +
  geom_histogram(binwidth = 1, center = 0, color = "grey") +
  geom_vline(data = species_ww_summary, mapping = aes(xintercept = pos_samples)) +
  facet_wrap(vars(species)) +
  scale_x_continuous(breaks = seq(0, 12, 2), name = "positive samples")
```

This plot compares the observed number of positive wastewater samples (vertical line) with the distribution generated from our model.
It looks like the model systematically overestimates the number of wastewater samples with viral reads (though not by a lot in any one case).
This may be a sign that $\phi_{ww}$ is too large: too little overdispersion leading to too few zero counts given a mean relative abundance.
It could also mean that the negative binomial model is slightly misspecified.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_total, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed total viral reads in wastewater (points) with the distribution predicted by our model.
The observations here are all very close to the posterior median.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_max, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed maximum viral reads in wastewater with the distribution from our model.

Here the model might be systematically slightly overestimating the maximum viral reads.
This cuts against the "too little overdispersion" hypothesis for the excess positive samples.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_sd, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed standard deviation of viral reads in wastewater with the distribution from our model.

Here the model might also be systematically slightly too high.

Taking these together, it could be that the negative binomial model can't simultaneously fit the number
of positive samples and the dispersion of reads because these are controlled by a single parameter.

I don't think this is a major issue for trusting the results.
The observed values are all well within the posterior predictive distributions
and it probably makes sense to be conservative with respect to the possibility of seeing anomalously large counts. 

# Results 

## Prevalence

```{r}
posteriors %>%
  ggplot(aes(p, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(name = "prevalence", limits = c(1e-4 / 4, 4e-1)) +
  scale_y_discrete(limits=rev(species_order))
```

```{r}
prev_posteriors <- left_join(
  swab_reads_complete %>%
    summarise(
      `+ pools` = sum(viral_reads > 0),
      `viral reads` = sum(viral_reads),
      .by = species,
    ),
  posteriors %>%
    summarise(
      q15 = quantile(p, .15),
      median = median(p),
      q85 = quantile(p, .85),
      `q85 / q15` = q85 / q15,
      .by = species,
    ),
  )

write_tsv(prev_posteriors, "prevalence_posteriors.tsv")

prev_posteriors %>%
  kable(digits = c(NA, 0, 0, 4, 4, 4, 2)) %>%
  column_spec(2:7, monospace = TRUE)
```
```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    mu_swab = first(mu_swab),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  summarise(
    median_mu_swab = median(mu_swab),
    median_p0 = median(mean_p0),
    .by = species,
  ) 
```

Let's compare these results to what we'd get with perfect sensitivity.
In that case, we can estimate the maximum likelihood $p$ and the Fisher information for $\log p$,
which gives an estimate of the posterior posterior mode and inverse variance of $\log p$ if we had a flat prior on $\log p$.

### Comparison with perfect sensitivity maximum likelihood estimate: 

```{r}
perfect_sensitivity <- swab_reads_complete %>%
    summarise(
      pos_pools = sum(viral_reads > 0),
      p = pos_pools / (sum(pool_size[viral_reads == 0]) + sum(pool_size[viral_reads >0]) / 2),
      sd_logp = 1 / sqrt(pos_pools),
      viral_reads = sum(viral_reads),
      .by = c(species, group),
      )

comparison <- posteriors %>%
  summarise(median(p), sd(log(p)), .by = c(species, group)) %>%
  left_join(
    perfect_sensitivity,
  ) %>%
  mutate(
    p_ratio = `median(p)` / p,
    sd_ratio = `sd(log(p))` / sd_logp,
    .keep = "unused",
  )

comparison %>%
  kable(digits = 1)

comparison %>%
  filter(group %in% c("Coronaviruses (SARS-CoV-2)", "Coronaviruses (seasonal)", "Rhinoviruses")) %>%
  summarise(
    n(),
    median(p_ratio),
    median(sd_ratio),
  )
```

```{r}
posteriors %>%
  ggplot(aes(x = p, y = species, fill = group)) +
  geom_vline(xintercept = plogis(mean_logit_p), color = "grey") +
  geom_vline(xintercept = plogis(mean_logit_p + sd_logit_p * c(1, -1)), color = "grey", linetype = "dashed") +
  geom_violin(draw_quantiles = c(0.15, .5, 0.85)) + 
  geom_point(aes(x = p), data = perfect_sensitivity, color = "white") +
  geom_errorbarh(aes(xmin = p / exp(sd_logp), xmax = p * exp(sd_logp)), data = perfect_sensitivity, color = "white") +
  scale_x_log10() +
  scale_y_discrete(limits=rev(species_order))
```

Here, the violins show posteriors with median, 15-th, and 85-th percentiles as black lines.
The white dots are the maximum likelihood estimates with error bars $\log \hat{p} \pm {\mathcal{I_{\log p}}}^{-1/2}$.
Solid grey line is the prior median and dashed grey lines are the prior 15th and 85th percentiles.

## RA_p(1%)

Finally, we can calculate the RA_p(1%) values - the relative abundance at 1% prevalence - which was a key goal of the original analysis:

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  ggplot(aes(ra01, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(name = "wastewater RA_p(1%)") +
  scale_y_discrete(limits=rev(species_order))
```

This plot shows the posterior distribution of RA_p(1%) for each virus species.
Vertical lines represent 15%, 50%, and 85% quantiles of the posterior.

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q05 = quantile(ra01, 0.05),
    q15 = quantile(ra01, 0.15),
    q25 = quantile(ra01, 0.25),
    median = median(ra01),
    q75 = quantile(ra01, 0.75),
    q85 = quantile(ra01, 0.85),
    q95 = quantile(ra01, 0.95),
    ratio_q85q15 = q85 / q15,
    gsd_sq = exp(2 * sd(log(ra01))),
    ratio_q95q05 = q95 / q05,
    .by = c(species, group),
  ) %>%
  mutate(
    across(
      -c(species, group, starts_with("ratio_"), gsd_sq),
      ~ format(.x, scientific = TRUE, digits = 2)
    ),
    across(
      c(starts_with("ratio_"), gsd_sq),
      ~ format(.x, digits = 2)
    )
  ) %>% 
  select(-group) %>%
  kable
```

This table provides quantiles of the RA_p(1%) estimates for each virus species, representing the uncertainty in these values.

```{r}
ra_posteriors <- posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q15 = quantile(ra01, .15),
    median = median(ra01),
    q85 = quantile(ra01, .85),
    `q85 / q15` = q85 / q15,
    .by = c(species)
  )

write_tsv(ra_posteriors, "ra_posteriors.tsv")

ra_posteriors %>% 
  mutate(
    across(
      c(q15, median, q85),
      ~ format(.x, scientific = TRUE, digits = 2)
    ),
    across(
      c(`q85 / q15`),
      ~ format(.x, digits = 2)
    )
  ) %>%
  kable(align = "lrrrr") %>%
  column_spec(2:5, monospace = TRUE)
```

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q15 = quantile(ra01, .15),
    median = median(ra01),
    q85 = quantile(ra01, .85),
    sd = exp(sd(log(ra01))),
    `q85 / q15` = q85 / q15,
    .by = c(species)
  ) %>%
  summarise(
    exp(sd(log(median))),
    median(sd),
  )
```

## RA swab / RA wastewater

```{r}
posteriors %>%
  mutate(ratio = mu_swab / mu_ww) %>%
  ggplot(aes(ratio, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(
    name = "swab RA / wastewater RA",
    limits = c(1e-2, 1e8),
    breaks = 10^seq(0, 6, 2)
  ) +
  scale_y_discrete(limits=rev(species_order))
```

This plot shows the ratio of swab relative abundance to wastewater relative abundance for a fixed prevalence.
Vertical lines represent quartiles of the posterior.

As a sanity check, let's calculate some two different estimates of this factor from the count data alone:
1. The ratio of relative abundances, summing reads over samples
2. The ratio of relative abundances, averaging RA over samples

```{r}
get_ra <- function(data){
  summarise(
    data,
    ra_by_read = sum(viral_reads) / sum(all_reads),
    ra_by_sample = mean(viral_reads / all_reads),
    .by = c(species, group)
  )
}

naive_ra_ratios <- left_join(
  get_ra(swab_reads_complete),
  get_ra(wastewater_reads_complete),
  by = join_by(species, group),
  suffix = c(".swab", ".ww"),
) %>%
  mutate(
    ratio_by_read = ra_by_read.swab / ra_by_read.ww,
    ratio_by_sample = ra_by_sample.swab / ra_by_sample.ww,
  )

posteriors %>%
  mutate(ratio = mu_swab / mu_ww) %>%
  ggplot(aes(ratio, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  geom_point(aes(ratio_by_read, species), naive_ra_ratios, shape = "x") +
  geom_point(aes(ratio_by_sample, species), naive_ra_ratios, shape = "o") +
  scale_x_log10(
    name = "swab RA / wastewater RA",
    limits = c(1e-2, 1e8),
    breaks = 10^seq(0, 6, 2)
  ) +
  scale_y_discrete(limits=rev(species_order))
```

I don't see a systematic bias here, which suggests that the priors aren't influencing the ratio too much.
