---
title: "Model checking"
format:
  html:
    code-fold: false
    toc: true
  pdf:
    toc: true
knitr:
  opts_chunk: 
    message: false
editor: source 
author: Dan Rice
date: 2025-05-27
date-modified: today
---

This document contains model checking for the swab/wastewater model.

# Setup

```{r}
#| label: setup

library(tidyverse)
library(knitr)
library(here)
library(kableExtra)

theme_set(theme_minimal())

source("../scripts/prepare_data.R")
species_order <- all_species$species
source("../scripts/define_model.R")

# Colors borrowed from https://github.com/JLSteenwyk/ggpubfigs
wong_eight <- c(
  "#E69F00",
  "#56B4E9",
  "#009E73",
  "#F0E442",
  "#0072B2",
  "#D55E00",
  "#CC79A7",
  "#000000"
)
options(
  ggplot2.discrete.colour = function()
    scale_colour_manual(values = wong_eight)
)
options(
  ggplot2.discrete.fill = function()
    scale_fill_manual(values = wong_eight)
)

# For axis labels
scientific <- function(x) {
  parse(text = sprintf("10^{%f}", log10(x)))
}

# Helper function for plotting priors
# If the x-axis is log10-transformed,transform the density function of x to
# the density of log10(x) by a Jacobian adjustment:
# dividing by $\frac{d log_{10}(x){dx} = \frac{1}{x \log 10}$
# so that the areas under the curve give probabilities correctly
prior_density <- function(x, density_function, log10x = FALSE){
  if (log10x){
    data <- tibble(
      x = x,
      density = density_function(x) * x * log(10), # Jacobian adjustment
    )
  } else {
    data <- tibble(
      x = x,
      density = density_function(x),
    )
  }
  geom_line(
    aes(x, density),
    data = data,
    color="grey",
    linetype="dashed",
    inherit.aes = FALSE
    )
}

```

## Data summaries

```{r}
species_swab_summary <- swab_reads_complete %>%
  summarise(
    pos_pools = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    .by = c(species, group),
  ) %>%
  arrange(species)
kable(species_swab_summary)
```

```{r}
species_ww_summary <- wastewater_reads_complete %>%
  summarise(
    pos_samples = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    .by = c(species, group),
  ) %>%
  arrange(species)
kable(species_ww_summary)
```

## Prior Distributions

```{r}
#| label: priors

prior_logit_p <- function(x) dnorm(x, mean_logit_p, sd_logit_p)
prior_logit_mu_swab <- function(x) dnorm(x, mean_logit_mu_swab, sd_logit_mu_swab)
prior_logit_mu_ww <- function(x) dnorm(x, mean_logit_mu_ww, sd_logit_mu_ww)
prior_phi_swab <- function(x) dgamma(x, shape_phi_swab, rate_phi_swab)
prior_phi_ww <- function(x) dgamma(x, shape_phi_ww, rate_phi_ww)
```

# Prior predictive checks

```{r}
#| label: prior-pred-check-swab
#| cache: true

n_prior_samples <- 1000
prior_samples <- tibble(
  rep = 1:n_prior_samples,
  p = plogis(rnorm(mean = mean_logit_p, sd = sd_logit_p, n = n_prior_samples)),
  mu_swab = plogis(rnorm(mean = mean_logit_mu_swab, sd = sd_logit_mu_swab, n = n_prior_samples)),
  phi_swab = rgamma(shape = shape_phi_swab, rate = rate_phi_swab, n = n_prior_samples),
) %>%
  cross_join(swab_metadata) %>%
  mutate(
    n_pos = map2_int(pool_size, p, ~ rbinom(1, .x, .y)),
    mu = n_pos * mu_swab * all_reads / pool_size,
    phi = n_pos * phi_swab,
    viral_reads = map2_int(mu, phi, ~ ifelse(.x > 0, rnbinom(
      1, size = .y, mu = .x
    ), 0)),
  )

prior_predictive_summary <- prior_samples %>%
  summarise(
    pos_swabs = sum(n_pos),
    pos_pools = sum(viral_reads > 1),
    exp_pos_swabs = sum(p * pool_size),
    p_no_pos_swabs = exp(-exp_pos_swabs),
    total_reads = sum(viral_reads),
    max_reads = max(viral_reads),
    .by = rep,
  ) 

prior_predictive_summary %>%
  summarise(
    frac_no_pos_swabs = mean(pos_swabs == 0),
    frac_no_pos_pools = mean(pos_pools == 0),
  )

prior_predictive_summary %>%
  ggplot(aes(x = pos_pools)) +
  geom_histogram(binwidth = 1) +
  geom_point(aes(y = 0), data = species_swab_summary, position = position_jitter()) 

prior_predictive_summary %>%
  filter(pos_pools > 0) %>%
  ggplot(aes(x = total_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_total, y = 0), data = species_swab_summary) +
  scale_x_continuous(transform = "log10", breaks = 10^seq(0:5))

prior_predictive_summary %>%
  filter(pos_pools > 0) %>%
  ggplot(aes(x = max_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_max, y = 0), data = species_swab_summary) +
  scale_x_continuous(transform = "log10", breaks = 10^seq(0:4))
```

```{r}
#| label: prior-pred-check-ww
#| cache: true

prior_samples_ww <- tibble(
  rep = 1:n_prior_samples,
  p = plogis(rnorm(mean = mean_logit_p, sd = sd_logit_p, n = n_prior_samples)),
  mu_ww = plogis(rnorm(mean = mean_logit_mu_ww, sd = sd_logit_mu_ww, n = n_prior_samples)),
  phi_ww = rgamma(shape = shape_phi_ww, rate = rate_phi_ww, n = n_prior_samples),
) %>%
  cross_join(swab_metadata) %>%
  mutate(
    mu = p * mu_ww * all_reads,
    viral_reads = map2_int(mu, phi_ww, ~ rnbinom(
      1, size = .y, mu = .x
    )),
  )

prior_predictive_summary_ww <- prior_samples_ww %>%
  summarise(
    pos_samples = sum(viral_reads > 1),
    total_reads = sum(viral_reads),
    max_reads = max(viral_reads),
    .by = rep,
  ) 

prior_predictive_summary_ww %>%
  summarise(mean(pos_samples == 0))

prior_predictive_summary_ww %>%
  ggplot(aes(x = pos_samples)) +
  geom_histogram(binwidth = 1) +
  geom_point(aes(y = 0), data = species_ww_summary, position = position_jitter()) 

prior_predictive_summary_ww %>%
  filter(pos_samples > 0) %>%
  ggplot(aes(x = max_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_max, y = 0), data = species_ww_summary) +
  scale_x_continuous(transform = "log10")

prior_predictive_summary_ww %>%
  filter(pos_samples > 0) %>%
  ggplot(aes(x = total_reads)) +
  geom_density() +
  geom_point(aes(x = viral_reads_total, y = 0), data = species_ww_summary, position = position_dodge()) +
  scale_x_continuous(transform = "log10")
```


## Model Diagnostics

After fitting the model, it's important to check diagnostics to ensure the model is reliable.
These diagnostics help us assess the quality of the model fits:

- Divergent transitions indicate potential issues with the model
- Maximum treedepth indicates how complex the posterior exploration was
- Rhat values close to 1 indicate good convergence

```{r}
diagnostics <- read_tsv(here(tables_dir, "model_diagnostics.tsv"))
kable(diagnostics)
```

No problems here.

# Posterior distributions

```{r}
posteriors <- read_tsv(here(tables_dir, "posteriors.tsv"))
```

## Marginal posterior distributions

Let's visualize the posterior distributions for our key parameters.
In the following plots, solid lines show the posterior distribution of prevalence for each virus species, compared to the prior distribution (dashed line).
Colors represent viral groups (SARS-CoV-2, seasonal Coronaviruses, Influenza, Mononegavirales, and Rhinoviruses).

For reference, here's a summary of the data:

```{r}
full_join(
  swab_reads_complete %>%
    summarise(
      `+ pools` = sum(viral_reads > 0),
      `swab reads` = sum(viral_reads),
      .by = species,
    ),
  wastewater_reads_complete %>%
    summarise(
      `+ ww samples` = sum(viral_reads > 0),
      `ww reads` = sum(viral_reads),
      .by = species,
    ),
  by = join_by(species)
  ) %>%
  kable
```

### Prevalence

```{r}
# Helper function for plotting priors
prior_density <- function(x, density_function, log10x = FALSE){
  if (log10x){
    data <- tibble(
      x = x,
      density = density_function(x) * x * log(10), # Jacobian adjustment
    )
  } else {
    data <- tibble(
      x = x,
      density = density_function(x),
    )
  }
  geom_line(
    aes(x, density),
    data = data,
    color="grey",
    linetype="dashed",
    inherit.aes = FALSE
    )
}
```

```{r}
posteriors %>%
  ggplot(aes(p, color=group)) +
  geom_density() +
  prior_density(plogis(seq(-12, 4, length.out=100)), compose(prior_logit_p, qlogis)) +
  scale_x_continuous(transform = "logit", limits = c(1e-5, 2e-1), breaks = 10^seq(-4, -1, 1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  labs(x = "prevalence", y = "") +
  theme(legend.position = "", axis.text.y = element_blank()) 
```

We can also plot the log density:

```{r}
posteriors %>%
  ggplot(aes(p, color=group)) +
  geom_density() +
  prior_density(plogis(seq(-12, 4, length.out=100)), compose(prior_logit_p, qlogis)) +
  scale_x_continuous(transform = "logit", limits = c(1e-5, 2e-1), breaks = 10^seq(-4, -1, 1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(
    transform = "log10",
    limits = c(1e-2, 1),
    breaks = 10^seq(0, -2),
    labels = scientific
  ) +
  labs(x = "prevalence", y = "") +
  theme(legend.position = "")
```

Comparison with perfect sensitivity maximum likelihood estimate: 

```{r}
perfect_sensitivity <- swab_reads_complete %>%
    summarise(
      pos_pools = sum(viral_reads > 0),
      p = pos_pools / (sum(pool_size[viral_reads == 0]) + sum(pool_size[viral_reads >0]) / 2),
      sd_logp = 1 / sqrt(pos_pools),
      viral_reads = sum(viral_reads),
      .by = c(species, group),
      )

posteriors %>%
  summarise(median(p), sd(log(p)), .by = c(species, group)) %>%
  left_join(
    perfect_sensitivity,
  ) %>%
  mutate(
    p_ratio = `median(p)` / p,
    sd_ratio = `sd(log(p))` / sd_logp,
    .keep = "unused",
  ) %>%
  kable(digits = 1)
```

```{r}
posteriors %>%
  ggplot(aes(x = p, y = species, fill = group)) +
  geom_vline(xintercept = plogis(mean_logit_p), color = "grey") +
  geom_vline(xintercept = plogis(mean_logit_p + sd_logit_p * c(1, -1)), color = "grey", linetype = "dashed") +
  geom_violin(draw_quantiles = c(0.15, .5, 0.85)) + 
  geom_point(aes(x = p), data = perfect_sensitivity, color = "white") +
  geom_errorbarh(aes(xmin = p / exp(sd_logp), xmax = p * exp(sd_logp)), data = perfect_sensitivity, color = "white") +
  scale_x_log10() +
  scale_y_discrete(limits=rev(species_order))
```

### Expected viral fractions

Posteriors on $\mu_{swab}$ (swab viral fraction with all positive swabs).

With linear y-scale:

```{r}
breaks <- 10^seq(-5, -1, 2)
posteriors %>%
  ggplot(aes(logit_mu_swab, color = group)) +
  geom_density() +
  prior_density(seq(-16, 2, length.out = 100), prior_logit_mu_swab) +
  scale_x_continuous(
    limits = c(-14, 2),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "", x = "mu_swab") +
  theme(legend.position = "", axis.text.y = element_blank()) 
```

And log y-scale:

```{r}
posteriors %>%
  ggplot(aes(logit_mu_swab, color = group)) +
  geom_density() +
  prior_density(seq(-16, 2, length.out = 100), prior_logit_mu_swab) +
  scale_x_continuous(
    limits = c(-14, 2),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "mu_swab") +
  theme(legend.position = "")
```

For species with multiple positive pools, (e.g. the Rhinoviruses), there is information to constrain the expected viral fraction in a positive swab.
With only one positive pool (e.g. H1N1, which has only one viral read), there is less, but still some information.

HCoV-229E is an outlier because it has only a single positive pool with 6008 viral reads.
This puts most of the weight on $\mu_{swab} > 0.1$, which would be higher if the prior weren't pulling it down.
I'm not sure whether this is a good effect: regularization, or a bias we'd like to eliminate.
It depends on whether we think it's likely that $\mu_{swab}$ could actually be this high, vs this being a very lucky sample.

Same plots but for $\mu_{ww}$:

```{r}
breaks <- 10^seq(-7, -3, 2)
posteriors %>%
  ggplot(aes(logit_mu_ww, color = group)) +
  geom_density() +
  prior_density(seq(-20, -5, length.out = 100), prior_logit_mu_ww) +
  scale_x_continuous(
    limits = c(-20, -5),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  scale_y_continuous(name = "") +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "", x = "mu_ww") +
  theme(legend.position = "", axis.text.y = element_blank()) 
```

```{r}
posteriors %>%
  ggplot(aes(logit_mu_ww, color = group)) +
  geom_density() +
  prior_density(seq(-20, -5, length.out = 100), prior_logit_mu_ww) +
  scale_x_continuous(
    limits = c(-20, -5),
    breaks = qlogis(breaks),
    labels = scientific(breaks)
  ) +
  facet_wrap(vars(species), ncol = 4) +
  scale_y_continuous(transform = "log10", limits = c(1e-3, 1)) +
  labs(y = "", x = "mu_ww") +
  theme(legend.position = "")
```

Here the posteriors are narrow compared to the prior and not far from the prior mode.

### Overdispersion

Posteriors for swab inverse-overdispersion $\phi_{swab}$.

```{r}
posteriors %>%
  ggplot(aes(phi_swab, color=group)) +
  geom_density() +
  prior_density(10^seq(-3, 3, length.out=100), prior_phi_swab, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1)) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "", axis.text.y = element_blank())
```

The posterior distributions are all very close to the prior,
suggesting that, as expected the data contains very little information about $\phi_{swab}$, at least relative to the prior.

There is a slight leftward shift for HCoV-OC43, which has the most positive pools (5).

Posteriors for wastewater inverse-overdispersion $\phi_{ww}$.

```{r}
posteriors %>%
  ggplot(aes(phi_ww, color=group)) +
  geom_density() +
  prior_density(10^seq(-2, 3, length.out=100), prior_phi_ww, log10x = TRUE) +
  scale_x_log10(limits = 10^c(-3,1), labels = scientific) +
  facet_wrap(vars(species), ncol = 4) +
  labs(y = "") +
  theme(legend.position = "", axis.text.y = element_blank())
```

The wastewater data has more non-zero counts and thus more data to infer dispersion.
I'm not sure I trust any of these though.

Generally, the species that have high predicted overdispersion relative to the prior (low $\phi$) are ones with fewer positive wastewater samples. 
That could be because there are more zero-read samples than expected given the counts in the non-zero-read samples.
Or it could be because the estimates are noisier with less data.

## Joint Posterior Distributions

We can also examine relationships between different parameters in the posterior.
In the following plots, each point is a sample from the joint posterior of two parameters.
The curves are an estimate of the conditional mean of the y-axis parameter given the x-axis parameter.

### Realtionships with prevalence

Prevalence vs $\mu_{swab}$ (note log x-scale not logit):

```{r}
posteriors %>%
  ggplot(aes(mu_swab, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(limits = c(1e-7, 1), labels = scientific) +
  scale_y_log10(limits = c(1e-5, 0.5), labels = scientific) +
  facet_wrap(~species)
```

It looks like:

- For small $\mu$, prevalence grows as $\mu$ decreases.
- For larger $\mu$, this relationship flattens.

This makes sense because low $\mu$ requires high prevalence to observe the virus since each positive swab is likely to go unobserved, while with high $\mu$ positive swabs are usually observed, so the prevalence is determined by the number of positive pools.

TODO: Add maximum likelihood calculation here.

```{r}
posteriors %>%
  ggplot(aes(phi_swab, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(limits = c(1e-3, 10), labels=scientific) +
  scale_y_log10(limits = c(1e-5, 0.5), labels = scientific) +
  facet_wrap(~species)
```

Smaller $\phi_{swab}$ implies a higher probability of failing to get viral reads a positive swab.
Thus, the inferred prevalence conditional on $\phi_{swab}$ is inversely related to $\phi_{swab}$.

```{r}
posteriors %>%
  summarise(
    corr_log_p_phi_swab = cor(log(p), log(phi_swab), method = "spearman"),
    corr_p_phi_swab = cor(p, phi_swab, method = "spearman"),
    corr_p_mu_ww = cor(log(p), log(mu_swab), method = "spearman"),
    .by = species,
  )
```

To encapsulate the relationship between $p$, $\mu_{swab}$, and $\phi_{swab}$,
we can approximate $\hat{p}$, the maximum likelihood $p$ conditional on known $\mu_{swab}$ and
$\phi_{swab}$.
(The approximation is in the the limit where most pools are expected to be negative.)

```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  ggplot(aes(p_hat, p)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() + 
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  scale_x_log10(limits = c(2e-3, 1)) +
  scale_y_log10() +
  facet_wrap(~species)
```
```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  summarise(
    corr_sp = cor(log(mean_p0), log(p), method = "spearman"),
    corr_pe = cor(log(mean_p0), log(p), method = "pearson"),
    .by = c(species),
  )
```

The approximate maximum likelihood and posterior $p$ qualitatively agree.

### Relationships with expected wastewater viral fraction

Prevalence: 

```{r}
posteriors %>%
  ggplot(aes(p, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```

```{r}
posteriors %>%
  summarise(
    cor_sp = cor(p, mu_ww, method = "spearman"),
    cor_log = cor(log(p), log(mu_ww), method = "pearson"),
    .by = species,
  )
```

As expected, there is a very tight relationship between prevalence and estimated wastewater viral fraction (per prevalence).
Often, strong correlations between parameters suggest that we should reparameterize the model so the parameters are more independent.
However, in this case, I think this is appropriate. We need a distinct $p$ parameter to make use of the discreteness of the swab pools.

The next three plots show the relationship between $\mu_{ww}$ and $\mu_{swab}$, $\phi_{swab}$, and $\phi_{ww}$:

```{r}
posteriors %>%
  ggplot(aes(mu_swab, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```


```{r}
posteriors %>%
  ggplot(aes(phi_swab, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scientific) +
  scale_y_log10(labels = scientific) +
  facet_wrap(~species)
```

```{r}
posteriors %>%
  ggplot(aes(phi_ww, mu_ww)) +
  geom_point(alpha = 0.05, size = 0.1) +
  geom_smooth() +
  scale_x_log10() + # label = scientific) +
  scale_y_log10(label = scientific) +
  facet_wrap(~species)
```

The relationships with $\mu_{swab}$ and $\phi_{swab}$ are mediated by $p$ and are not surprising.

The relationship with $\phi_{ww}$ suggests that the posteriors are uncorrelated,
which is encouraging because we don't have any real information about this parameter.

# Posterior Predictive Checks

## Swab data

To validate our model, we can use posterior predictive checks - generating new data from the fitted model and comparing it to the observed data.

```{r}
#| label: swab-ppd
#| cache: true

posterior_predictive_swab <- posteriors %>%
  mutate(rep = 1:n(), .by = species) %>%
  cross_join(swab_metadata) %>%
  mutate(
    n_pos = map2_int(pool_size, p, ~ rbinom(1, .x, .y)),
    mu = n_pos * mu_swab * all_reads / pool_size,
    phi = n_pos * phi_swab,
    viral_reads = map2_int(mu, phi, ~ ifelse(.x > 0, rnbinom(
      1, size = .y, mu = .x
    ), 0)),
    false_negative = n_pos > 0 & viral_reads == 0,
  )

posterior_predictive_swab_summary <- posterior_predictive_swab %>%
  summarise(
    pos_pools = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    false_negative_pools = sum(false_negative),
    across(c(p, mu_swab, phi_swab), first),
    .by = c(species, group, rep),
  )
```


```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = pos_pools, fill = group, after_stat(density))) +
  geom_histogram(binwidth = 1, center = 0, color = "grey") +
  geom_vline(data = species_swab_summary, mapping = aes(xintercept=pos_pools)) +
  facet_wrap(vars(species)) +
  scale_x_continuous(breaks = seq(0, 12, 2), name = "positive pools")
```

This plot compares the observed number of positive pools (vertical line) with the distribution of positive pools generated from our model.
For every virus, the observed number of positive pools is common in the simulations.

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_total, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_continuous(
    transform = "log10",
    breaks = 10^seq(0, 4),
    name = "total viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed total viral reads (points) with the distribution from our model.
Vertical bars show quartiles

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_max, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_continuous(
    # transform = "pseudo_log",
    # breaks = c(0, 10^seq(0, 4)),
    transform = "log10",
    breaks = 10^seq(0, 4),
    name = "largest viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed maximum viral reads with the distribution from our model.

```{r}
posterior_predictive_swab_summary %>%
  ggplot(aes(x = viral_reads_sd, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_swab_summary) +
  scale_x_log10(
    breaks = 10^seq(0, 4),
    name = "stdev viral read count (conditional on > 0)",
    oob = scales::oob_keep,
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed standard deviation of viral reads with the distribution from our model.
Violins show only the simulations with at least one viral read.
Vertical bars show quartiles (conditional on at least one read).

Together, these plots suggest that the model is capturing the swab data reasonably well.

## Wastewater data

Now we'll perform similar posterior predictive checks for the wastewater data:

```{r}
#| label: wastewater-ppd
#| cache: true

posterior_predictive_ww <- posteriors %>%
  mutate(rep = 1:n(), .by=species) %>%
  cross_join(wastewater_metadata) %>%
  mutate(
    mu = p * mu_ww * all_reads,
    viral_reads = map2_int(mu, phi_ww, ~ ifelse(.x > 0, rnbinom(1, size = .y, mu = .x), 0)), 
  )

posterior_predictive_ww_summary <- posterior_predictive_ww %>%
  summarise(
    pos_samples = sum(viral_reads > 0),
    viral_reads_total = sum(viral_reads),
    viral_reads_max = max(viral_reads),
    viral_reads_sd = sd(viral_reads),
    across(c(p, mu_ww, phi_ww), first),
    viral_reads_expected = p * mu_ww * sum(all_reads),
    .by = c(species, group, rep),
  )
```

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = pos_samples, fill = group, after_stat(density))) +
  geom_histogram(binwidth = 1, center = 0, color = "grey") +
  geom_vline(data = species_ww_summary, mapping = aes(xintercept = pos_samples)) +
  facet_wrap(vars(species)) +
  scale_x_continuous(breaks = seq(0, 12, 2), name = "positive samples")
```

This plot compares the observed number of positive wastewater samples (vertical line) with the distribution generated from our model.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_total, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed total viral reads in wastewater (points) with the distribution predicted by our model.
The observations here are all very close to the posterior median.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_max, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed maximum viral reads in wastewater with the distribution from our model.

```{r}
posterior_predictive_ww_summary %>%
  ggplot(aes(x = viral_reads_sd, y = species, fill = group)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_point(data = species_ww_summary) +
  scale_x_continuous(
    transform = "pseudo_log",
    breaks = c(0, 10^seq(0, 4)),
  ) +
  scale_y_discrete(limits = rev(species_order))
```

This plot compares the observed standard deviation of viral reads in wastewater with the distribution from our model.

These posterior predictev plots all look good.

# Results 

## Prevalence

```{r}
posteriors %>%
  ggplot(aes(p, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(name = "prevalence", limits = c(1e-4 / 4, 4e-1)) +
  scale_y_discrete(limits=rev(species_order))
```

```{r}
prev_posteriors <- left_join(
  swab_reads_complete %>%
    summarise(
      `+ pools` = sum(viral_reads > 0),
      `viral reads` = sum(viral_reads),
      .by = species,
    ),
  posteriors %>%
    summarise(
      q15 = quantile(p, .15),
      median = median(p),
      q85 = quantile(p, .85),
      `q85 / q15` = q85 / q15,
      .by = species,
    ),
  )

prev_posteriors %>%
  kable(digits = c(NA, 0, 0, 4, 4, 4, 2)) %>%
  column_spec(2:7, monospace = TRUE)
```
```{r}
posteriors %>%
  mutate(posterior_sample = seq(nrow(.))) %>%
  left_join(swab_reads_complete, relationship = "many-to-many", by = join_by(species, group)) %>%
  mutate(
    lambda = mu_swab * all_reads / pool_size,
    p0 = (phi_swab / (lambda + phi_swab)) ^ phi_swab,
  ) %>%
  summarise(
    p = first(p),
    mu_swab = first(mu_swab),
    pos_pools = sum(viral_reads > 0),
    p_hat = pos_pools / sum(pool_size[viral_reads == 0] * (1 - p0[viral_reads == 0])),
    mean_p0 = mean(p0),
    .by = c(species, group, posterior_sample),
  ) %>%
  summarise(
    median_mu_swab = median(mu_swab),
    median_p0 = median(mean_p0),
    .by = species,
  ) 
```

### Comparison with perfect sensitivity maximum likelihood estimate: 

Let's compare these results to what we'd get with perfect sensitivity.
In that case, we can estimate the maximum likelihood $p$ and the Fisher information for $\log p$,
which gives an estimate of the posterior posterior mode and inverse variance of $\log p$ if we had a flat prior on $\log p$.

```{r}
perfect_sensitivity <- swab_reads_complete %>%
    summarise(
      pos_pools = sum(viral_reads > 0),
      p = pos_pools / (sum(pool_size[viral_reads == 0]) + sum(pool_size[viral_reads >0]) / 2),
      sd_logp = 1 / sqrt(pos_pools),
      viral_reads = sum(viral_reads),
      .by = c(species, group),
      )

comparison <- posteriors %>%
  summarise(median(p), sd(log(p)), .by = c(species, group)) %>%
  left_join(
    perfect_sensitivity,
  ) %>%
  mutate(
    p_ratio = `median(p)` / p,
    sd_ratio = `sd(log(p))` / sd_logp,
    .keep = "unused",
  )

comparison %>%
  kable(digits = 1)

comparison %>%
  filter(group %in% c("Coronaviruses (SARS-CoV-2)", "Coronaviruses (seasonal)", "Rhinoviruses")) %>%
  summarise(
    n(),
    median(p_ratio),
    median(sd_ratio),
  )
```

```{r}
posteriors %>%
  ggplot(aes(x = p, y = species, fill = group)) +
  geom_vline(xintercept = plogis(mean_logit_p), color = "grey") +
  geom_vline(xintercept = plogis(mean_logit_p + sd_logit_p * c(1, -1)), color = "grey", linetype = "dashed") +
  geom_violin(draw_quantiles = c(0.15, .5, 0.85)) + 
  geom_point(aes(x = p), data = perfect_sensitivity, color = "white") +
  geom_errorbarh(aes(xmin = p / exp(sd_logp), xmax = p * exp(sd_logp)), data = perfect_sensitivity, color = "white") +
  scale_x_log10() +
  scale_y_discrete(limits=rev(species_order))
```

Here, the violins show posteriors with median, 15-th, and 85-th percentiles as black lines.
The white dots are the maximum likelihood estimates with error bars $\log \hat{p} \pm {\mathcal{I_{\log p}}}^{-1/2}$.
Solid grey line is the prior median and dashed grey lines are the prior 15th and 85th percentiles.

## RA_p(1%)

Finally, we can calculate the RA_p(1%) values - the relative abundance at 1% prevalence - which was a key goal of the original analysis:

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  ggplot(aes(ra01, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(name = "wastewater RA_p(1%)") +
  scale_y_discrete(limits=rev(species_order))
```

This plot shows the posterior distribution of RA_p(1%) for each virus species.
Vertical lines represent 15%, 50%, and 85% quantiles of the posterior.

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q05 = quantile(ra01, 0.05),
    q15 = quantile(ra01, 0.15),
    q25 = quantile(ra01, 0.25),
    median = median(ra01),
    q75 = quantile(ra01, 0.75),
    q85 = quantile(ra01, 0.85),
    q95 = quantile(ra01, 0.95),
    ratio_q85q15 = q85 / q15,
    gsd_sq = exp(2 * sd(log(ra01))),
    ratio_q95q05 = q95 / q05,
    .by = c(species, group),
  ) %>%
  mutate(
    across(
      -c(species, group, starts_with("ratio_"), gsd_sq),
      ~ format(.x, scientific = TRUE, digits = 2)
    ),
    across(
      c(starts_with("ratio_"), gsd_sq),
      ~ format(.x, digits = 2)
    )
  ) %>% 
  select(-group) %>%
  kable
```

This table provides quantiles of the RA_p(1%) estimates for each virus species, representing the uncertainty in these values.

```{r}
ra_posteriors <- posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q15 = quantile(ra01, .15),
    median = median(ra01),
    q85 = quantile(ra01, .85),
    `q85 / q15` = q85 / q15,
    .by = c(species)
  )

ra_posteriors %>% 
  mutate(
    across(
      c(q15, median, q85),
      ~ format(.x, scientific = TRUE, digits = 2)
    ),
    across(
      c(`q85 / q15`),
      ~ format(.x, digits = 2)
    )
  ) %>%
  kable(align = "lrrrr") %>%
  column_spec(2:5, monospace = TRUE)
```

```{r}
posteriors %>%
  mutate(ra01 = mu_ww * 0.01) %>%
  summarise(
    q15 = quantile(ra01, .15),
    median = median(ra01),
    q85 = quantile(ra01, .85),
    sd = exp(sd(log(ra01))),
    `q85 / q15` = q85 / q15,
    .by = c(species)
  ) %>%
  summarise(
    exp(sd(log(median))),
    median(sd),
  )
```

## RA swab / RA wastewater

```{r}
posteriors %>%
  mutate(ratio = mu_swab / mu_ww) %>%
  ggplot(aes(ratio, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  scale_x_log10(
    name = "swab RA / wastewater RA",
    limits = c(1e-2, 1e8),
    breaks = 10^seq(0, 6, 2)
  ) +
  scale_y_discrete(limits=rev(species_order))
```

This plot shows the ratio of swab relative abundance to wastewater relative abundance for a fixed prevalence.
Vertical lines represent quartiles of the posterior.

As a sanity check, let's calculate some two different estimates of this factor from the count data alone:
1. The ratio of relative abundances, summing reads over samples
2. The ratio of relative abundances, averaging RA over samples

```{r}
get_ra <- function(data){
  summarise(
    data,
    ra_by_read = sum(viral_reads) / sum(all_reads),
    ra_by_sample = mean(viral_reads / all_reads),
    .by = c(species, group)
  )
}

naive_ra_ratios <- left_join(
  get_ra(swab_reads_complete),
  get_ra(wastewater_reads_complete),
  by = join_by(species, group),
  suffix = c(".swab", ".ww"),
) %>%
  mutate(
    ratio_by_read = ra_by_read.swab / ra_by_read.ww,
    ratio_by_sample = ra_by_sample.swab / ra_by_sample.ww,
  )

posteriors %>%
  mutate(ratio = mu_swab / mu_ww) %>%
  ggplot(aes(ratio, species, fill=group)) +
  geom_violin(draw_quantiles = c(0.15, 0.5, 0.85)) +
  geom_point(aes(ratio_by_read, species), naive_ra_ratios, shape = "x") +
  geom_point(aes(ratio_by_sample, species), naive_ra_ratios, shape = "o") +
  scale_x_log10(
    name = "swab RA / wastewater RA",
    limits = c(1e-2, 1e8),
    breaks = 10^seq(0, 6, 2)
  ) +
  scale_y_discrete(limits=rev(species_order))
```

I don't see a systematic bias here, which suggests that the priors aren't influencing the ratio too much.
